### 2022-04-20 manuscript/manuscript.Rmd
prepared the preliminary results and created a manuscript.

---
title: "Detecting skill level and mental template in biface morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Radu Iovita^[Department of Anthropology, New York University, New York, NY, USA; Department of Early Prehistory and Quaternary Ecology, Eberhard Karls University of Tübingen, Tübingen, Germany; iovita@nyu.edu]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain challenging to interpret. \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

# Introduction

Stone tools

# Materials and Methods

## Boxgrove biface collection

To Radu: I cannot find any context information of this sample from your 2011 JHE paper [@iovita2011]. Could you please write something here?

## Experimental biface collection

The biface experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given.

In this experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas. Subsequently, the 17 participants in the experimental group were assessed after every ten hours of the cumulative learning period, where each of them was requested to produce a handaxe for expert knapper's review, leading to the compilation of a data set composing 9 assessements in total. It should be also noted that 6 out of 17 participants dropped out of the research before the final assessment due to personal reasons. To detect the effect of training on skill level and mental template, we reorganized our assessment classification scheme and combined it into three broader categories, namely pre-training (assessment 1), early training (assessment 2-5), and late training (assessment 6-9), which helps increase the sample size of the measured intervals. A more detailed experimental protocol can be assessed in one of our published papers [@pargeter2019].

## Lithic Analysis

To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples prepared by R. I. [@iovita2011], which include both the front and lateral views of a given specimen. First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting. This is then followed by the batch conversion of color photographs to a black-and-white binary format. Subsequently, we cropped the silhouettes of bifaces one by one using the Quick Selection Tool in Adobe Photoshop. The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette. The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements). Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006]. The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al. [-@pargeter2019].

```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the biface measurement protocal using Image J (after Pargeter et al. 2019: Figure 5)."}
knitr::include_graphics(here::here("figure/Fig1.png"))
```

## Statistical Analyses

Given the number of variables involved in this study, we used the principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most commonly used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020]. To detect the effect of training on novices' performance as compared with archaeological samples and biface made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github ([\<https://github.com/Raylc/PaST-pilot\>](https://github.com/Raylc/PaST-pilot){.uri}).

# Results

## Principal Component analysis

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather decent explained variance ratio to avoid overfitting. We then decided to focus on and further interpreted the implications of these first two components based on their relationships between variables (**Table** \@ref(tab:tab1)). The first principle component (PC1) indicates the overall biface thickness as it is positively correlated with all thickness measurements while negatively correlated with all other measurements. That being said, a higher PC1 value indicates a thicker biface, and vice versa. The second principle component (PC2) tracks the elongation and pointedness based on its positive relationship with maximum length and bottom width/thickness. As PC2 increases, a biface will be generally longer and more pointed since its bottom part will be bulkier.

```{r Screeplot, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principle components."}
knitr::include_graphics(here::here("figure/Scree plot.png"))
```

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/variable_loadings_pc12.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principle components") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

```

A closer look at the principal component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) yields the clustering of different groups of bifaces. The majority of Boxgrove bifaces occupy an area featuring negative values of both PC1 and PC2. The expert group is similar to the Boxgrove group in PC1, while the former has a relatively higher PC2 value than the latter on average. The group of novice displays the highest level of variability, however, it is rather pronounced that most bifaces made by novices have a positive PC1 value that is different from both the groups of Boxgrove and experts.

```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of bifaces from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
knitr::include_graphics(here::here("figure/General PCA1.png"))
```

## Effects of training

We extracted the PC1 and PC2 values of individual bifaces and compared them between different groups. More specifically, the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert and the one between Early training and Late training stages, which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is **not** statistically significant. However, a rather surprising result here is that the mean PC2 value difference between the Pre-training group and Boxgrove is also **not** statistically significant.

```{r PCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A between-group comparison of PC1 values."}
knitr::include_graphics(here::here("figure/PC1 comparison.png"))
```

```{r PCA2, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A between-group comparison of PC2 values."}
knitr::include_graphics(here::here("figure/PC2 comparison.png"))
```

# Discussion

The most salient finding of our exploratory study is that

# Conclusions

Our exploratory study produced findings with both theoretical and methodological implications.

# CRediT authorship contribution statement

**Cheng Liu**: Data curation, Formal analysis, Investigation, Methodology, Visualization, Writing -- original draft, Writing -- review & editing. **Radu Iovita**: Resources, Writing -- original draft, Writing -- review & editing. **Dietrich Stout**: Conceptualization, Investigation, Resources, Funding acquisition, Supervision, Writing -- original draft, Writing -- review & editing. **Justin Pargeter**: Conceptualization, Investigation, Methodology, Supervision, Writing -- original draft, Writing -- review & editing.

# Declaration of competing interest

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

# Acknowledgements

This work was supported by funding from the National Science Foundation of the USA (grants SMA-1328567 & DRL-1631563), the John Templeton Foundation (grant 47994), and the Emory University Research Council. The handaxe knapping skill acquisition experiment involved in this study was approved by Emory University's Internal Review Board (IRB study no: 00067237).

# References

### 2022-04-20 code/Boxgrove_Iovita.R

@@ -14,6 +14,7 @@ require(MASS)
 require(cowplot)
 library(sjstats)
 library(pwr)
+library(patchwork)
 ############################## 
 # Set working directory and source files
 ##############################
@@ -185,6 +186,11 @@ fviz_pca_ind(bepca1, habillage= boxgrove_experiment_data_shapes1$group, # color 
              ellipse.type = "convex",
              legend.title = "Groups",
              label="none")
+ggsave("General PCA1.png", path="figure", dpi = 600)
+
+# Cumulative scree plot
+fviz_eig(bepca1, addlabels=TRUE, hjust = -0.1)
+ggsave("Scree plot.png", path="figure", dpi = 600)
 
 # Compute standard deviation of each principal component
 
@@ -194,18 +200,6 @@ std_dev<- boxgrove_experiment_pca$sdev
 
 pr_var <- std_dev^2
 prop_varex <- pr_var/sum(pr_var)
-
-# Scree plot
-
-pc_scree_plot<-plot(prop_varex, xlab = "Principal Component",
-                    ylab = "Proportion of Variance Explained",
-                    type = "b")
-
-# Cumulative scree plot
-
-pc_cumulative_plot<-plot(cumsum(prop_varex), xlab = "Principal Component",
-                         ylab = "Cumulative Proportion of Variance Explained",
-                         type = "b")
 
 # Graph of variables
 
@@ -229,6 +223,10 @@ eig.val <- get_eigenvalue(boxgrove_experiment_pca)
 
 res.var <- get_pca_var(boxgrove_experiment_pca)
 variable_loadings<-data.frame(res.var$coord)         # Coordinates
+variable_loadings_pc12 <- subset(variable_loadings, select=c("Dim.1", "Dim.2"))
+write.csv(variable_loadings_pc12,"data\\variable_loadings_pc12.csv", row.names = TRUE)
+
+
 
 # Results for individuals
 
@@ -240,27 +238,41 @@ Indiv_handaxe_scores<-data.frame(res.ind$coord)        # Coordinates
 boxgrove_experiment_data_shapes$PC1<-Indiv_handaxe_scores$Dim.1
 boxgrove_experiment_data_shapes$PC2<-Indiv_handaxe_scores$Dim.2
 
+# plotting the pc1 and pc2 by assessment stage
+boxgrove_experiment_data_shapes2 <- boxgrove_experiment_data_shapes%>%
+  filter(assessment != "10" & assessment != "11")
+
[…]

### 2022-04-21 manuscript/manuscript.Rmd
manuscript outline added.

@@
 
-Stone tools
+Handaxe is a well-studied and well-published topic, but some aspects of it still remain elusive.
+More specifically, the sources of variation of handaxe morphology can be very complex, but skill level and mental template are two major factors.
 
-# Materials and Methods
+Skill level: short literature review
+
+Mental template: short literature review
+
+Here we have two interconnected research questions based on a reference sample generated from our 90-hour handaxe skill acquisition experiment: 1) Can skill level and mental template (or "aesthetic preference") be effeciently detected from biface morphometric data?
+2) What is the effect of training on these two aspects?
+
+# Materials and methods
 
@@
 The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014].
-The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total.
+The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total.
 Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given.
@@
 
-## Lithic Analysis
+## Lithic analysis
 
@@
 
-## Statistical Analyses
+## Statistical analyses
 
@@
 
-## Principal Component analysis
+## Principal component analysis
 
@@
 We then decided to focus on and further interpreted the implications of these first two components based on their relationships between variables (**Table** \@ref(tab:tab1)).
-The first principle component (PC1) indicates the overall biface thickness as it is positively correlated with all thickness measurements while negatively correlated with all other measurements.
+The first principal component (PC1) indicates the overall biface thickness as it is positively correlated with all thickness measurements while negatively correlated with all other measurements.
 That being said, a higher PC1 value indicates a thicker biface, and vice versa.
-The second principle component (PC2) tracks the elongation and pointedness based on its positive relationship with maximum length and bottom width/thickness.
+The second principal component (PC2) tracks the elongation and pointedness based on its positive relationship with maximum length and bottom width/thickness.
 As PC2 increases, a biface will be generally longer and more pointed since its bottom part will be bulkier.
 
-```{r Screeplot, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principle components."}
+```{r Screeplot, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
 knitr::include_graphics(here::here("figure/Scree plot.png"))
@@
 df1<-read.csv(here::here("data/variable_loadings_pc12.csv"), fileEncoding = 'UTF-8-BOM')
-kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principle components") %>%
+kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components") %>%
   kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
@@
 
-The most salient finding of our exploratory study is that
+PC1 (thickness) is a robust indicator of skill level as it is shared by modern expert knapper and boxgrove foragers, while PC2 (elongation and pointedness) reflects more of personal/community-level aesthetic choices.
+However, we do not intend to construct a false binary framework and put these two factors as disconnected and opposite concepts.
+
+Training is important in terms of its immediate effects (pre-training vs.
+post-training) in both skill level and mental template, however, 90 hours of training is still not enough for novice to reach the reach the skill level as reflected in expert knappers or Late Acheulean toolmaker.
+
+The pre-training group is similar to the Boxgrove group in PC2 because of the raw material restriction in experimental set up, which is typically flat flakes?
+Novice before training cannot effectively reduce the nodule to a pointed shape, making their products assembling Boxgrove in this dimension.
+
+Experimental archaeology has a huge potential in similar topics.
 
@@
 
-Our exploratory study produced findings with both theoretical and methodological implications.
+Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with thickness while the latter is reflected in elongation and pointedness; 2) Training has an effect, but 90 hours of training is still not enough for novice to reach the level of expertise.
 

### 2022-04-25 manuscript/manuscript.Rmd
added the context of boxgrove assemblage.

@@
 
-To Radu: I cannot find any context information of this sample from your 2011 JHE paper [@iovita2011]. Could you please write something here?
+The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposit [@roberts1998; @pope2020].
+This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominins' subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
+In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high knapping skill level reflected in this manufacture of these handaxes.
+As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019].
+To identify the morphological manifestation of knappers' dexterity in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
 
@@
 
-Given the number of variables involved in this study, we used the principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most commonly used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020].
+Given the number of variables involved in this study, we used the principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most commonly used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b].
 To detect the effect of training on novices' performance as compared with archaeological samples and biface made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance.
-This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github ([\<https://github.com/Raylc/PaST-pilot\>](https://github.com/Raylc/PaST-pilot){.uri}).
+This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github (<https://github.com/Raylc/PaST-pilot>).
 
@@
 
+Maybe also something about how this study demonstrate the potential of archaeological data reuse and its contempoary relevance [information age+pandemic+ethical research]?
+
 # Conclusions

### 2022-04-28 code/Boxgrove_Iovita.R
added the delta weight analysis

@@ -355,3 +355,34 @@ ggstatsplot::ggbetweenstats(
 )
 ggplot2::ggsave("PC2 comparison.png", path="figure", dpi = 600)
 
+
+
+data1 <- read.csv("data/Experiment/GM worksheet.csv")
+data2 <- read.csv("data/Experiment/Core Weights.csv")
+total <- inner_join(data1, data2, by="Starting_Weight")
+total <- total %>%
+  mutate(delta_Weight= Starting_Weight-Weight)
+total_grouped <-total %>%
+  mutate(assessment_stage=ifelse(assessment %in% c("1"),"Pre-training",
+                                 ifelse(assessment %in% c("2","3","4","5"),"Early training",
+                                        ifelse(assessment %in% c("6","7","8","9"),"Late training", "NA"))))
+total_grouped$assessment_stage <- relevel(total_grouped$assessment_stage, 'Late training')
+
+p <- ggstatsplot::ggbetweenstats(
+  data  = total_grouped,
+  x     = assessment_stage,
+  y     = delta_Weight,
+  ggsignif.args = list(textsize = 1.5, tip_length = 0.01),
+  ggplot.component = list(scale_x_discrete(limits=c("Late training", "Early training", "Pre-training"))),
+  title = "A between-group comparison of delta weight"
+)
+ggplot2::ggsave("delta weight comparison1.png", path="figure", dpi = 600)
+
+ggstatsplot::extract_stats(p)
+
+
+aov_dw<-aov(delta_Weight~assessment_stage,data=total_grouped)
+summary(aov_dw)
+anova_stats(aov_dw)
+
+TukeyHSD(aov_dw)

### 2022-05-19 manuscript/manuscript.Rmd
remove apel's data and add some text in discussion.

@@
 - Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
+- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
 - Radu Iovita^[Department of Anthropology, New York University, New York, NY, USA; Department of Early Prehistory and Quaternary Ecology, Eberhard Karls University of Tübingen, Tübingen, Germany; iovita@nyu.edu]
@@
 
-The pre-training group is similar to the Boxgrove group in PC2 because of the raw material restriction in experimental set up, which is typically flat flakes?
-Novice before training cannot effectively reduce the nodule to a pointed shape, making their products assembling Boxgrove in this dimension.
+The pre-training group is similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to a desired form (**Figure** \@ref(fig:comparison)).
+That being said, if the given nodules already possess a oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged.
+This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group display the lowest value (**Figure** \@ref(fig:deltaweight)).
+On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by foragers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360].
+These characteristics have been clearly displayed in a recent attempt of slow motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>).
+As such, behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.
+
+```{r comparison, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the process"}
+knitr::include_graphics(here::here("figure/CAPTURE.PNG"))
+```
+
+```{r deltaweight, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A comparison of the delta weight between the pre-training, early training, late training, and the expert group."}
+knitr::include_graphics(here::here("figure/delta weight comparison3.png"))
+```
 
@@
 **Cheng Liu**: Data curation, Formal analysis, Investigation, Methodology, Visualization, Writing -- original draft, Writing -- review & editing.
-**Radu Iovita**: Resources, Writing -- original draft, Writing -- review & editing.
+**Nada Khreisheh**: Investigation, Writing -- review & editing.
+**Radu Iovita**: Resources, Writing -- review & editing.
 **Dietrich Stout**: Conceptualization, Investigation, Resources, Funding acquisition, Supervision, Writing -- original draft, Writing -- review & editing.

### 2022-05-19 code/Boxgrove_Iovita.R

@@ -26,14 +26,22 @@ getwd()
 # Add necessary datasets
 ##############################
 
-# Size/shape data
+# Boxgrove Size/shape data
 plan_filenames<-list.files("data/Boxgrove_Iovita/Measurements/plan", pattern="*.txt")
 profile_filenames<-list.files("data/Boxgrove_Iovita/Measurements/profile", pattern="*.txt")
 
-# Size/area data
+# Boxgrove Size/area data
 metric_filenames<-list.files("data/Boxgrove_Iovita/Measurements/metric", pattern = "*.csv")
 
-#experimental handaxe data
+# Experiment nodule Size/shape data
+nodule_plan_filenames<-list.files("data/Experiment/Nodule_Pretraining/Measurements/plan", pattern="*.txt")
+nodule_profile_filenames<-list.files("data/Experiment/Nodule_Pretraining/Measurements/profile", pattern="*.txt")
+
+# Experiment nodule Size/area data
+nodule_metric_filenames<-list.files("data/Experiment/Nodule_Pretraining/Measurements/metric", pattern = "*.csv")
+
+
+# Experimental handaxe data
 experimental_data<-read.csv("data/Experiment/experimental_handaxes.csv")
 
 experimental_data<-experimental_data %>%
@@ -89,12 +97,47 @@ merge_shape_data_function<-function(plan_names,profile_names,plan_pathway,profil
   return(merged_shape_data)
 }
 
+# Merge plans and profiles of pre-training nodules
+merge_shape_data_function1<-function(plan_names,profile_names,plan_pathway,profile_pathway) {
+  
+  # combine txt files for each recording set
+  handaxe_plan_measurements<-do.call(rbind,lapply(plan_names,open_read,plan_pathway))
+  handaxe_profile_measurements<-do.call(rbind,lapply(profile_names,open_read,profile_pathway))
+  
+  # clean and reshape plan data 
+  handaxe_plan_measurements_reshape<-handaxe_plan_measurements %>%
+    mutate(variable=recode(handaxe_plan_measurements$variable, Width = "width"),
+           measurement_point = paste(variable, measurement_point, sep="_"),
+           shape_width_mm=measurement*10) %>%
+    dplyr::select(-c(variable,measurement)) %>%
+    dcast(individual ~ measurement_point,value.var="shape_width_mm")
+  
+  # clean and reshape profile data 
[…]

### 2022-05-20 manuscript/manuscript.Rmd
added another paragraph in discussion.

@@
 Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given.
+As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through Neolithics.com into flat blanks of similar size and shape for training and assessments.
 
@@
 
-Training is important in terms of its immediate effects (pre-training vs.
-post-training) in both skill level and mental template, however, 90 hours of training is still not enough for novice to reach the reach the skill level as reflected in expert knappers or Late Acheulean toolmaker.
+In terms of our second research question, this study shows that training do have an immediate intervention effects (pre-training vs.
+post-training) in both PC1 (skill level) and PC2 (mental template).
+Nontheless, once the training has been initiated, its effects across different assessments on both dimensions are rather unconsipicous.
+This finding corroborates what has been suggested in Pargeter et al.
+[-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, considering the massive social support involved in the experiment set up including the direct and deliberate peadagogy and the simplified raw material procurement and preparation procedures.
+This follow-up project further adds the samples produced by Late Acheulean toolmaker as a new benchmark to deepen our understanding on this issue.
+It is noteworthy that how constrained is the Boxgrove assemblage morphological variation as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied group.
+Some potential explainations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].
 
@@
 
-Maybe also something about how this study demonstrate the potential of archaeological data reuse and its contempoary relevance [information age+pandemic+ethical research]?
+Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions.
+In this paper, the main source of archaeological data is a collection of photos produced and curated by one of our co-authors (R.
+Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas.
+Given the irreversible nature of archaeological excavation, the digitized data, be it text, pictures, or videos, often become the sole evidence that are available for certain research questions.
+Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].
+Among many reasons of preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].
+As stated in the method section, we addressed this issue through sharing the raw data and the code for generating the derived data on an open-access repository.
+Another major and legitimate concern of archaeological data resue is their quality.
+In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to more detailed technological analysis of stone artifacts, however, our paper shows that finding the approriate research questions given the data available is key to revealing new novel insights into the studied topic.
+Moreover, we believe that this type of research has a strong contempoarary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].
 

### 2022-06-03 manuscript/manuscript.Rmd
added the first paragraph of the introduction.

@@
 
-Handaxe is a well-studied and well-published topic, but some aspects of it still remain elusive.
-More specifically, the sources of variation of handaxe morphology can be very complex, but skill level and mental template are two major factors.
+The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019].
+Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
+More specifically, a complex suite of interconnecting factors have been identified to contribute the great variation of biface morphology, encompassing raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.
 
-Skill level: short literature review
+Skill level: short literature review. Symmetry, thinning,
 
@@
 
-Given the number of variables involved in this study, we used the principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most commonly used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b].
+Given the number of variables involved in this study, we used the principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most commonly used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017].
 To detect the effect of training on novices' performance as compared with archaeological samples and biface made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance.

### 2022-06-06 manuscript/manuscript.Rmd
added the paragraph of mental template.

@@
 - Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
-abstract: "Stone tools provide key evidence of human cognitive evolution but remain challenging to interpret.
-\\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
+abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic still remain elusive.
+Among many factors intentified to contribute to the considerable variation of biface morphology, skill level and mental template have been frequently quoted and discussed.
+In this paper, we \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
 fontsize: 11pt
@@
 Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors have been identified to contribute the great variation of biface morphology, encompassing raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.
+More specifically, a complex suite of interconnecting factors have been identified to contribute the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.
+
+In its classical definition, mental template is the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45].
+This concept essentially lies in the very foundation of the cultural historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial temporal framework.
+Early researchers, whether explicitly or implicitly, often endorse this conceptual framework and actively applies in the typological analysis of biface at regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
+Combined with the production of large flakes, the emergence of mental template (or "imposed form") has been recongnized as two major technological innovations of the Acheulean compared with the Oldowan [@isaac1986].
+For a decade or so, this concept has been less frequently used, since it was criticized for its normative and static assumption [@lyman2004], ignorance of other competing factors such as raw material constraint [@white1995], and the lacking of rigorous studies of its corresponding cognitive processes.
+Until very recently, several researchers have actively addressed these critiques and reconceuptualized the mental template in the study of biface morphology.
+Hutchence and Scott [-@hutchence2021: 675], for example, leveraged the theory of "community of practice" [@wenger1998] and argued that "the consistency in handaxe shape found at sites like Boxgrove is a consequence of the emergent actions of individual knappers being simultaneously constrained by the imposition of social norms." Further, raw material is often treated as a crucial variable to be tested in the very beginning of a research design focusing on mental template [@garcía-medrano2019].
+In terms of the cognitive mechanisms behind mental template, Ho and collegues [-@ho2022] recently developed a series of nevigation experiments demonstrating the externalization of the planning process to simple geometric representations, which has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors.
+In short, when exercized with proper caution, mental template still has its value in our study of biface morphological variation.
 
@@
 
-Mental template: short literature review
-
-Here we have two interconnected research questions based on a reference sample generated from our 90-hour handaxe skill acquisition experiment: 1) Can skill level and mental template (or "aesthetic preference") be effeciently detected from biface morphometric data?
-2) What is the effect of training on these two aspects?
+Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers.
+Since the novice biface collection is generated from an 90-hour skill acquisition experiment, we also have the precious opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
+As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental template be efficiently detected from biface morphometric data?
+2) How does training affect novices' performance on these two aspects?
 
@@
 The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements).
-Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006].
+Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006], which also at least alleviate the effect of resharpening process to some extent.
 The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al.

### 2022-06-06 manuscript/manuscript.Rmd
finished the very first draft.

--- 2022-06-03 manuscript/manuscript.Rmd
@@
 - Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
-abstract: "Stone tools provide key evidence of human cognitive evolution but remain challenging to interpret.
+abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.
+Among many factors identified to contribute to the considerable variation of biface morphology, skill level and mental template have been frequently quoted and discussed.
+In this paper, we compared the experimental replicas made by modern expert and novice knappers with the Boxgrove handaxe assemblage.
+Through the multivariate analysis of morphometric data, our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to thickness while the latter is connected with elongation and pointedness.
+Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the level of modern expert requires more training time than what is permitted in the experimental setting.
+Our study demonstrated the potential of experimental archaeology and digitized secondary data in revealing new insights from old archaeological assemblages.
 \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
@@
 Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors have been identified to contribute the great variation of biface morphology, encompassing raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.
+More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.
 
-Skill level: short literature review. Symmetry, thinning,
+In its classical definition, mental template is the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45].
+This concept essentially lies in the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework.
+Early researchers, whether explicitly or implicitly, often endorse this conceptual framework and actively applies in the typological analysis of biface at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
+Combined with the production of large flakes, the emergence of mental template (or "imposed form") has been recognized as two major technological innovations of the Acheulean compared with the Oldowan [@isaac1986].
+For a decade or so, this concept has been less frequently used, since it was criticized for its normative and static assumption [@lyman2004], ignorance of other competing factors such as raw material constraint [@white1995], and the lacking of rigorous studies of its corresponding cognitive processes.
+Until very recently, several researchers have actively addressed these critiques and reconceptualized the mental template in the study of biface morphology.
+Hutchence and Scott [-@hutchence2021: 675], for example, leveraged the theory of "community of practice" [@wenger1998] and argued that "the consistency in handaxe shape found at sites like Boxgrove is a consequence of the emergent actions of individual knappers being simultaneously constrained by the imposition of social norms." Furthermore, raw material is often treated as a crucial variable to be tested at the very beginning of a research design focusing on mental template [@garcía-medrano2019].
+In terms of the cognitive mechanisms behind mental template, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations, which has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors.
+In short, when exercised with proper caution, mental template still has its value in our study of biface morphological variation.
 
-Mental template: short literature review
+Building upon the concept of mental template, one possible way of defining skill is the capacity of realizing the former using the resources available [@roux1995: 66].
+This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989].
+It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993].
+As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem".
+This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of mental template in that they both refute the idea that technology is simply an internal program expressed by mind and prefer a dynamic approach emphasizing the interaction between the perception and action.
+The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners and it is thus possible to evaluate the skill level reflected in the knapping products based on time spent on practice [@stout2002; @roux1995].
+When contextual information is less readily available as in many technological systems, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
+In the context of biface technology, in addition to measurements that can be almost applied in any lithic technological system such as raw material and hammer choices, platform preparation, as well as hinges, symmetry [@hodgson2015; @hutchence2019] and thinning [@caruana2020; @stout2014] have been frequently quoted as reliable and distinctive indicators of the skill level as supported very several experimental studies.
+These two have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].
 
-Here we have two interconnected research questions based on a reference sample generated from our 90-hour handaxe skill acquisition experiment: 1) Can skill level and mental template (or "aesthetic preference") be effeciently detected from biface morphometric data?
-2) What is the effect of training on these two aspects?
+Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers.
+Since the novice biface collection is generated from a 90-hour skill acquisition experiment, we also have the precious opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
+As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental template be efficiently detected from biface morphometric data?
+2) How does training affect novices' performance in these two aspects?
 
@@
 Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given.
-As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through Neolithics.com into flat blanks of similar size and shape for training and assessments.
+As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through [Neolithics.com](https://neolithics.com/) into flat blanks of similar size and shape for training and assessments.
+The mechanical properties of these raw materials are comparable to the ones used in Boxgrove in that they are both fine-grained and highly predictable in fracturing process.
 
@@
 
-PC1 (thickness) is a robust indicator of skill level as it is shared by modern expert knapper and boxgrove foragers, while PC2 (elongation and pointedness) reflects more of personal/community-level aesthetic choices.
-However, we do not intend to construct a false binary framework and put these two factors as disconnected and opposite concepts.
+In accordiance with the existing literature on biface knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (thickness) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove foragers.
+Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the biface into several pieces, serving the purpose of achieving the desired convexity and/or volume.
+This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013].
+Experimental studies have also shown that the thinning stage of biface produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394].
+This also reflects the majority of samples in both our expert and novice experimental assemblages.
+In the skill acquisition experiments, novice knappers have been explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment.
+On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], corresponding to the cases of replicas produced by Bruce Bradley and a few novices in our study.
 
-In terms of our second research question, this study shows that training do have an immediate intervention effects (pre-training vs.
+Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis.
+This divergence of group-level aesthetic choices can be best explained under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in biface analysis[-@hutchence2021].
+The most common form of learning in the experiment occurred in the group condition, where the instructor taught multiple novices at the same time and novices have the chance to also communicated and learned from their peers.
+Unfortunately, the biface data from the instructor (N.
+Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley).
+This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.
+
+In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs.
 post-training) in both PC1 (skill level) and PC2 (mental template).
-Nontheless, once the training has been initiated, its effects across different assessments on both dimensions are rather unconsipicous.
+Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather unconsipicous.
 This finding corroborates what has been suggested in Pargeter et al.
-[-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, considering the massive social support involved in the experiment set up including the direct and deliberate peadagogy and the simplified raw material procurement and preparation procedures.
-This follow-up project further adds the samples produced by Late Acheulean toolmaker as a new benchmark to deepen our understanding on this issue.
-It is noteworthy that how constrained is the Boxgrove assemblage morphological variation as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied group.
-Some potential explainations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].
+[-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures.
+This follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue.
+It is noteworthy how constrained is the Boxgrove assemblage morphological variation as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied group.
+Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].
 
-The pre-training group is similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to a desired form (**Figure** \@ref(fig:comparison)).
-That being said, if the given nodules already possess a oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged.
-This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group display the lowest value (**Figure** \@ref(fig:deltaweight)).
+The pre-training group is similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
+That being said, if the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged.
+This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:deltaweight)).
 On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by foragers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360].
-These characteristics have been clearly displayed in a recent attempt of slow motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>).
+These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>).
 As such, behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.
@@
 
-Experimental archaeology has a huge potential in similar topics.
-
 Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions.
@@
 Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas.
-Given the irreversible nature of archaeological excavation, the digitized data, be it text, pictures, or videos, often become the sole evidence that are available for certain research questions.
+Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions.
 Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].
-Among many reasons of preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].
-As stated in the method section, we addressed this issue through sharing the raw data and the code for generating the derived data on an open-access repository.
-Another major and legitimate concern of archaeological data resue is their quality.
-In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to more detailed technological analysis of stone artifacts, however, our paper shows that finding the approriate research questions given the data available is key to revealing new novel insights into the studied topic.
-Moreover, we believe that this type of research has a strong contempoarary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].
+Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].
+As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository.
+Another major and legitimate concern of archaeological data reuse is their quality.
+In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic.
+Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].
 
@@
 
-Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with thickness while the latter is reflected in elongation and pointedness; 2) Training has an effect, but 90 hours of training is still not enough for novice to reach the level of expertise.
+Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with thickness while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of sharing a common mental template, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove foragers.
+However, it should be noted that it is not our intention to construct a false binary framework and put these two factors as disconnected and opposite concepts.
+In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of bifaces as well as their implications for the biological and cultural evolution of the hominin lineages.
 

### 2022-06-20 manuscript/manuscript.Rmd
revised based on Justin's comments.

@@
 ---
-title: "Detecting skill level and mental template in biface morphology: Archaeological and experimental insights"
+title: "Detecting skill level and mental templates in Late Acheulean biface morphology: Archaeological and experimental insights"
 author: 
@@
 abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.
-Among many factors identified to contribute to the considerable variation of biface morphology, skill level and mental template have been frequently quoted and discussed.
-In this paper, we compared the experimental replicas made by modern expert and novice knappers with the Boxgrove handaxe assemblage.
-Through the multivariate analysis of morphometric data, our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to thickness while the latter is connected with elongation and pointedness.
-Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the level of modern expert requires more training time than what is permitted in the experimental setting.
-Our study demonstrated the potential of experimental archaeology and digitized secondary data in revealing new insights from old archaeological assemblages.
+Among many factors identified to contribute to the considerable variation of biface morphology, skill level and mental templates have been frequently cited.
+Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving twenty-six naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
+We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
+Through the principal component analysis of morphometric data derived from images, our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
+Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern expert requires more training time than was permitted in this extensive and long-running training program.
+Our study demonstrated the potential of experimental archaeology and digital photographs in revealing new insights from old archaeological assemblages.
 \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
@@
 Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.
+More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021].
+From this extensive list, skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.
 
@@
 This concept essentially lies in the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework.
-Early researchers, whether explicitly or implicitly, often endorse this conceptual framework and actively applies in the typological analysis of biface at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
-Combined with the production of large flakes, the emergence of mental template (or "imposed form") has been recognized as two major technological innovations of the Acheulean compared with the Oldowan [@isaac1986].
-For a decade or so, this concept has been less frequently used, since it was criticized for its normative and static assumption [@lyman2004], ignorance of other competing factors such as raw material constraint [@white1995], and the lacking of rigorous studies of its corresponding cognitive processes.
+Early researchers, whether explicitly or implicitly, often endorse this conceptual framework and actively applies in the typological analysis of bifaces at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
+Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as two major technological innovations of the Acheulean compared with the Oldowan [@isaac1986].
+For a decade or so, this concept has been less frequently used, since it was criticized for its normative and static assumption [@lyman2004], ignorance of other competing factors such as raw material constraints [@white1995], and the lack of rigorous studies of its corresponding cognitive processes.
 Until very recently, several researchers have actively addressed these critiques and reconceptualized the mental template in the study of biface morphology.
-Hutchence and Scott [-@hutchence2021: 675], for example, leveraged the theory of "community of practice" [@wenger1998] and argued that "the consistency in handaxe shape found at sites like Boxgrove is a consequence of the emergent actions of individual knappers being simultaneously constrained by the imposition of social norms." Furthermore, raw material is often treated as a crucial variable to be tested at the very beginning of a research design focusing on mental template [@garcía-medrano2019].
+Hutchence and Scott [-@hutchence2021: 675], for example, leveraged the theory of "community of practice" [@wenger1998] and argued that "the consistency in handaxe shape found at sites like Boxgrove is a consequence of the emergent actions of individual knappers being simultaneously constrained by the imposition of social norms." Furthermore, raw material is often treated as a crucial variable to be controlled at the very beginning of a research design focusing on mental template.
+This is best exemplified by an experimental study of García-Medrano et al.
+[-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in archaeological context in composition, size, and shape.
 In terms of the cognitive mechanisms behind mental template, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations, which has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors.
-In short, when exercised with proper caution, mental template still has its value in our study of biface morphological variation.
+In short, when exercised with proper caution, mental template still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness and elongation, among other things.
 
-Building upon the concept of mental template, one possible way of defining skill is the capacity of realizing the former using the resources available [@roux1995: 66].
+Building upon the concept of mental template, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
 This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989].
@@
 As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem".
-This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of mental template in that they both refute the idea that technology is simply an internal program expressed by mind and prefer a dynamic approach emphasizing the interaction between the perception and action.
-The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners and it is thus possible to evaluate the skill level reflected in the knapping products based on time spent on practice [@stout2002; @roux1995].
-When contextual information is less readily available as in many technological systems, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
-In the context of biface technology, in addition to measurements that can be almost applied in any lithic technological system such as raw material and hammer choices, platform preparation, as well as hinges, symmetry [@hodgson2015; @hutchence2019] and thinning [@caruana2020; @stout2014] have been frequently quoted as reliable and distinctive indicators of the skill level as supported very several experimental studies.
-These two have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].
+This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action.
+The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners.
+It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].
+When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
+In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of biface technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
+These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].
 
@@
 This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominins' subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
-In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high knapping skill level reflected in this manufacture of these handaxes.
+In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high dexerity reflected in their manufacture.
 As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019].
 To identify the morphological manifestation of knappers' dexterity in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
+The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].
 
@@
 The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014].
+These handaxes were made for previous research projects, which similarly aimed to approximate 'Late Acheulean' handaxes explicitly comparable to the Boxgrove assemblage [@faisal2010; @stout2011a; @stout2014].
 The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total.
@@
 In this experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas.
-Subsequently, the 17 participants in the experimental group were assessed after every ten hours of the cumulative learning period, where each of them was requested to produce a handaxe for expert knapper's review, leading to the compilation of a data set composing 9 assessements in total.
+Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage.
+She was present at all training sessions to provide help and instruction to participants.
+All training occurred under controlled conditions at the outdoor knapping area of Emory's Paleolithic Technology Lab, with knapping tools and raw materials provided.
+All participants were instructed in basic knapping techniques including how to select appropriate percussors, initiate flaking on a nodule, maintain the correct flaking gestures and angles, prepare flake platforms, visualize outcomes, deal with raw material imperfections, and correct mistakes.
+Handaxe-specific instruction included establishment and maintenance of a bifacial plane, cross-sectional thinning, and overall shaping.
+The training emphasized both aspects of handaxe making technical skill (the importance of producing thin pieces with centered edges) as well as mental template related markers (symmetrical edges).
+
+Subsequently, the 17 participants in the experimental group were assessed after every ten hours of the cumulative learning period, where each of them was requested to produce a handaxe for expert knapper's (N.
+Khreisheh) review, leading to the compilation of a data set composing 9 assessments in total.
 It should be also noted that 6 out of 17 participants dropped out of the research before the final assessment due to personal reasons.
@@
 Subsequently, we cropped the silhouettes of bifaces one by one using the Quick Selection Tool in Adobe Photoshop.
-The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette.
+The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom ImageJ script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette.
 The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements).
@@
 
-Given the number of variables involved in this study, we used the principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most commonly used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017].
+Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017].
 To detect the effect of training on novices' performance as compared with archaeological samples and biface made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance.
@@
 We then decided to focus on and further interpreted the implications of these first two components based on their relationships between variables (**Table** \@ref(tab:tab1)).
-The first principal component (PC1) indicates the overall biface thickness as it is positively correlated with all thickness measurements while negatively correlated with all other measurements.
+The first principal component (PC1) indicates the overall cross-sectional thickness as it is positively correlated with all thickness measurements while negatively correlated with all other measurements.
 That being said, a higher PC1 value indicates a thicker biface, and vice versa.
@@
 More specifically, the novice group was divided into three sub-groups based on their training stages as specified in the method section.
-As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert and the one between Early training and Late training stages, which at least partially confirms our visual observation of the general PCA scatter plot.
-Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is **not** statistically significant.
-However, a rather surprising result here is that the mean PC2 value difference between the Pre-training group and Boxgrove is also **not** statistically significant.
+As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot.
+Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is **not** statistically significant ($t=0.333$, $p>0.05$).
+However, a rather surprising result here is that the mean PC2 value difference between the Pre-training group and Boxgrove is also **not** statistically significant ($t=-0.818$, $p>0.05$).
 
@@
 
-In accordiance with the existing literature on biface knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (thickness) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove foragers.
+In accordance with the existing literature on biface knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers.
 Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the biface into several pieces, serving the purpose of achieving the desired convexity and/or volume.
-This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013].
+This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013].
 Experimental studies have also shown that the thinning stage of biface produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394].
@@
 The pre-training group is similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
-That being said, if the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged.
+If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged.
 This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:deltaweight)).
-On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by foragers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360].
+On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360].
 These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>).
@@
 
-Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with thickness while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of sharing a common mental template, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove foragers.
+Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of sharing a common mental template, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove foragers.
 However, it should be noted that it is not our intention to construct a false binary framework and put these two factors as disconnected and opposite concepts.

### 2022-06-20 code/Boxgrove_Iovita.R

@@ -424,7 +424,7 @@ ggplot(data = PC2_summary,aes(x=assessment_stage,y=mean,fill=assessment_stage))+
 ggsave("PC2 score_New.png", path="figure", dpi = 600)
 
 
-ggstatsplot::ggbetweenstats(
+p1<-ggstatsplot::ggbetweenstats(
   data  = boxgrove_experiment_data_shapes,
   x     = assessment_stage,
   y     = PC1,
@@ -432,8 +432,9 @@ ggstatsplot::ggbetweenstats(
   title = "A between-group comparison of PC1 values"
 )
 ggplot2::ggsave("PC1 comparison.png", path="figure", dpi = 600)
+ggstatsplot::extract_stats(p1)
 
-ggstatsplot::ggbetweenstats(
+p2<-ggstatsplot::ggbetweenstats(
   data  = boxgrove_experiment_data_shapes,
   x     = assessment_stage,
   y     = PC2,
@@ -441,6 +442,7 @@ ggstatsplot::ggbetweenstats(
   title = "A between-group comparison of PC2 values"
 )
 ggplot2::ggsave("PC2 comparison.png", path="figure", dpi = 600)
+ggstatsplot::extract_stats(p2)
 
 
 

### 2022-07-24 manuscript/manuscript.Rmd
revised the introduction and added the correct figures.

@@
 - Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
-- Radu Iovita^[Department of Anthropology, New York University, New York, NY, USA; Department of Early Prehistory and Quaternary Ecology, Eberhard Karls University of Tübingen, Tübingen, Germany; iovita@nyu.edu]
 - Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
@@
 Through the principal component analysis of morphometric data derived from images, our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern expert requires more training time than was permitted in this extensive and long-running training program.
+Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
 Our study demonstrated the potential of experimental archaeology and digital photographs in revealing new insights from old archaeological assemblages.
@@
 The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019].
-Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
+Despite the recurrent narrative emphasizing the homogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
 More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021].
@@
 
-In its classical definition, mental template is the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45].
-This concept essentially lies in the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework.
-Early researchers, whether explicitly or implicitly, often endorse this conceptual framework and actively applies in the typological analysis of bifaces at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
-Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as two major technological innovations of the Acheulean compared with the Oldowan [@isaac1986].
-For a decade or so, this concept has been less frequently used, since it was criticized for its normative and static assumption [@lyman2004], ignorance of other competing factors such as raw material constraints [@white1995], and the lack of rigorous studies of its corresponding cognitive processes.
-Until very recently, several researchers have actively addressed these critiques and reconceptualized the mental template in the study of biface morphology.
-Hutchence and Scott [-@hutchence2021: 675], for example, leveraged the theory of "community of practice" [@wenger1998] and argued that "the consistency in handaxe shape found at sites like Boxgrove is a consequence of the emergent actions of individual knappers being simultaneously constrained by the imposition of social norms." Furthermore, raw material is often treated as a crucial variable to be controlled at the very beginning of a research design focusing on mental template.
+In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45].
+This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework.
+Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applies it in the typological analysis of bifaces at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
+Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986].
+For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes.
+To avoid the historical baggage associated with this controversial term, some researchers have also developed alternative frameworks such as "design imperatives" purely derived from ergonomic principles, which refers to a set of minimum features shared by all handaxes including glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
+
+Until very recently, several researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of biface morphology.
+Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
+They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms.
+This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al.
+[-@mcnabb2004], which tries to provide a more balanced perspective incorporating both individual agency and social learning.
+Furthermore, raw material is often treated as a crucial variable to be controlled at the very beginning of a research design focusing on mental templates.
 This is best exemplified by an experimental study of García-Medrano et al.
 [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in archaeological context in composition, size, and shape.
-In terms of the cognitive mechanisms behind mental template, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations, which has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors.
-In short, when exercised with proper caution, mental template still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness and elongation, among other things.
+In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources., Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains.
+In short, when exercised with proper caution, the concept of mental template still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness and elongation, among other things.
 
-Building upon the concept of mental template, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
+Following the reconceptualization of the concept of mental template, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
 This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989].
@@
 
-Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers.
+Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers.
 Since the novice biface collection is generated from a 90-hour skill acquisition experiment, we also have the precious opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
@@
 
-To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples prepared by R.
-I.
-[@iovita2011], which include both the front and lateral views of a given specimen.
+To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples [@iovita2011], which include both the front and lateral views of a given specimen.
 First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting.
@@
 knitr::include_graphics(here::here("figure/General PCA1.png"))
+```
+
+In addition, visual inspection of the principle component scatter plot (Fig.
+) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups.
+To test this,  we  conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)).
+Across all three groups, a negative correlation has been displayed between the PC1 and PC2 values, although this trend is not statistically significant (r=-0.41, p= 0.24) in the expert group, probably because of its small sample size.
+
+```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132)."}
+knitr::include_graphics(here::here("figure/PC correlation.png"))
 ```
@@
 Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions.
-In this paper, the main source of archaeological data is a collection of photos produced and curated by one of our co-authors (R.
-Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas.
+In this paper, the main source of archaeological data is a collection of photos produced more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas.
 Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions.
@@
 **Nada Khreisheh**: Investigation, Writing -- review & editing.
-**Radu Iovita**: Resources, Writing -- review & editing.
 **Dietrich Stout**: Conceptualization, Investigation, Resources, Funding acquisition, Supervision, Writing -- original draft, Writing -- review & editing.
@@
 The handaxe knapping skill acquisition experiment involved in this study was approved by Emory University's Internal Review Board (IRB study no: 00067237).
+We would also like to thank Radu Iovita for providing us access to the digital photographs of the Boxgrove handaxe assemblage.
 

### 2022-07-24 code/Boxgrove_Iovita.R

@@ -15,12 +15,7 @@ require(cowplot)
 library(sjstats)
 library(pwr)
 library(patchwork)
-############################## 
-# Set working directory and source files
-##############################
 
-# Set to source of shape and size data txt files
-getwd()
 
 ############################## 
 # Add necessary datasets
@@ -252,7 +247,6 @@ ggplot(data = boxgrove_experiment_data_shapes2) +
   geom_point()+
   geom_smooth()
 
-
 # plot pc 1,2 by group
 
 boxgrove_experiment_data_scaled<-boxgrove_experiment_data_scaled %>%
@@ -316,25 +310,6 @@ anova_stats(aov_PC2)
 
 TukeyHSD(aov_PC2)
 
-ggplot(data = PC1_summary,aes(x=assessment_stage,y=mean,fill=assessment_stage))+
-  geom_bar(stat="identity")+
-  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2)+
-  scale_color_viridis_d()+
-  theme(legend.position="none")+
-  scale_x_discrete(name="")+
-  scale_y_continuous(name="Average PC 1 score")+
-  theme(text = element_text(size=27))
-ggsave("PC1 score_New.png", path="figure", dpi = 600)
-
-ggplot(data = PC2_summary,aes(x=assessment_stage,y=mean,fill=assessment_stage))+
-  geom_bar(stat="identity")+
-  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2)+
-  scale_color_viridis_d()+
-  theme(legend.position="none")+
-  scale_x_discrete(name="")+
-  scale_y_continuous(name="Average PC 2 score")+
-  theme(text = element_text(size=27))
-ggsave("PC2 score_New.png", path="figure", dpi = 600)
 
 
 p1<-ggstatsplot::ggbetweenstats(
@@ -358,45 +333,19 @@ ggplot2::ggsave("PC2 comparison.png", path="figure", dpi = 600)
 ggstatsplot::extract_stats(p2)
 
 
+write.csv(boxgrove_experiment_data_shapes,"data/Experiment/boxgrove_experiment_data_shapes.csv", row.names = FALSE)
 
-
-
-
-
-
-
-
-ggstatsplot::ggscatterstats(
-  data  = boxgrove_experiment_data_shapes,
-  x     = PC1,
-  y     = PC2,
-  xlab  = "PC1",
-  ylab  = "PC2",
[…]

### 2022-07-25 manuscript/manuscript.Rmd
removed the old figures and revised the conclusion.

@@
 
-Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of sharing a common mental template, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove foragers.
+Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of sharing a common mental template, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers.
+At a bigger theoretical level it breaks down the distinction between social learning of design targets and individual learning of the skills needed to achieve them.
 However, it should be noted that it is not our intention to construct a false binary framework and put these two factors as disconnected and opposite concepts.
+To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
 In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of bifaces as well as their implications for the biological and cultural evolution of the hominin lineages.

### 2022-07-26 manuscript/manuscript.Rmd
updated the whole manuscript.

@@
 
-In this experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas.
+In this experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas from our expert sample.
 Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage.
@@
 The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements).
-Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006], which also at least alleviate the effect of resharpening process to some extent.
+Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006].
+This allometric scaling procedure at least alleviate the effect of resharpening process on differential handaxe size to some extent, given that Shipton and Clarkson's [@shipton2015] experiment has suggested that reduction intensity does not have a strong impact on the shape of handaxes.
 The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al.
@@
 
-Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather decent explained variance ratio to avoid overfitting.
-We then decided to focus on and further interpreted the implications of these first two components based on their relationships between variables (**Table** \@ref(tab:tab1)).
-The first principal component (PC1) indicates the overall cross-sectional thickness as it is positively correlated with all thickness measurements while negatively correlated with all other measurements.
-That being said, a higher PC1 value indicates a thicker biface, and vice versa.
-The second principal component (PC2) tracks the elongation and pointedness based on its positive relationship with maximum length and bottom width/thickness.
-As PC2 increases, a biface will be generally longer and more pointed since its bottom part will be bulkier.
+Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable explained variance ratio to avoid overfitting.
+Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness.
+It is positively correlated with all thickness measurements while negatively correlated with all other measurements.
+A higher PC1 value thus indicates a thicker biface, and vice versa.
+The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness.
+As PC2 increases, a biface will be relatively longer and more convergent from the broad base to the tip..
+Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.
 
@@
 Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is **not** statistically significant ($t=0.333$, $p>0.05$).
-However, a rather surprising result here is that the mean PC2 value difference between the Pre-training group and Boxgrove is also **not** statistically significant ($t=-0.818$, $p>0.05$).
+An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also **not** statistically significant ($t=-0.818$, $p>0.05$).
 
@@
 
+Our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
+Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
 In accordance with the existing literature on biface knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers.
@@
 In the skill acquisition experiments, novice knappers have been explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment.
-On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], corresponding to the cases of replicas produced by Bruce Bradley and a few novices in our study.
+On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], corresponding to the cases of replicas produced by Bruce Bradley.
 
 Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis.
-This divergence of group-level aesthetic choices can be best explained under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in biface analysis[-@hutchence2021].
-The most common form of learning in the experiment occurred in the group condition, where the instructor taught multiple novices at the same time and novices have the chance to also communicated and learned from their peers.
+Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by  Shipton and White [-@shipton2020] as well as Garcia-Medrano et al.
+[-@garcía-medrano2019].
+This phenomenon could be regarded as a divergence of group-level aesthetic choices and could be best explained under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in biface analysis[-@hutchence2021].
+The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.
+Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions.
 Unfortunately, the biface data from the instructor (N.
@@
 
+The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form.
+Our results (Fig.) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value).
+In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface.
+Consequently, it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed.
+Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers.
+This also provides an alternative explanation to the social transmission of form for the experimental convergence of on pointed forms.
+In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.
+
 In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs.
 post-training) in both PC1 (skill level) and PC2 (mental template).
-Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather unconsipicous.
+Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather inconsipicous.
 This finding corroborates what has been suggested in Pargeter et al.
-[-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures.
+[-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures.
 This follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue.
-It is noteworthy how constrained is the Boxgrove assemblage morphological variation as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied group.
-Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].
+It is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
+Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice[@milks2019].
 
@@
 This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:deltaweight)).
+It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape.
+Experts generally try to achieve handaxe forms while removing as little mass as possible (i.e.
+making as big a handaxe as possible from the nodule).
 On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360].
@@
 Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions.
-In this paper, the main source of archaeological data is a collection of photos produced more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas.
+In this paper, the main source of archaeological data is a collection of photos produced and curated by one of our co-authors (R.
+Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original artifacts.
 Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions.

### 2022-08-10 manuscript/manuscript.Rmd
finished another round of revision of the paper.

@@
 Among many factors identified to contribute to the considerable variation of biface morphology, skill level and mental templates have been frequently cited.
-Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving twenty-six naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
+Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for minimally 0 hours (controlled group) and maximally 90 hours in Late Acheulean style handaxe production and three expert knappers.
 We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
@@
 More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021].
-From this extensive list, skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.
+From this extensive list, skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
+Of particular attention here are the experimental works conducted by Stout et al.
+[-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al.
+[-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage.
+Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage through comparing it with a unique experimental collection.
 
@@
 Until very recently, several researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of biface morphology.
-Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
+Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
 They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms.
@@
 [-@mcnabb2004], which tries to provide a more balanced perspective incorporating both individual agency and social learning.
-Furthermore, raw material is often treated as a crucial variable to be controlled at the very beginning of a research design focusing on mental templates.
+As for the critique towards confounding factors explaining morphological variability, raw material is often treated as a crucial variable to be controlled at the very beginning of a research design focusing on mental templates.
 This is best exemplified by an experimental study of García-Medrano et al.
 [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in archaeological context in composition, size, and shape.
-In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources., Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains.
+In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources.
+Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains.
 In short, when exercised with proper caution, the concept of mental template still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness and elongation, among other things.
 
-Following the reconceptualization of the concept of mental template, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
+Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
 This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989].
@@
 Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006].
-This allometric scaling procedure at least alleviate the effect of resharpening process on differential handaxe size to some extent, given that Shipton and Clarkson's [@shipton2015] experiment has suggested that reduction intensity does not have a strong impact on the shape of handaxes.
+This allometric scaling procedure controls for size variation which may come from initial blanks and/or reduction intensity (shaping/resharpening).
+Notably, Shipton and Clarkson [-@shipton2015] previously found that reduction intensity does not have a strong impact on the shape of handaxes.
 The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al.
@@
 [-@garcía-medrano2019].
-This phenomenon could be regarded as a divergence of group-level aesthetic choices and could be best explained under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in biface analysis[-@hutchence2021].
+This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in biface analysis[-@hutchence2021].
 The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.
@@
 Consequently, it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed.
+It is possible that such constraints help to explain the convergence of our novice knappers on similar shapes to those prefered by modern expert knappers, however, this clearly does not explain the design target at Boxgrove.
 Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers.
@@
 
-Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of sharing a common mental template, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers.
-At a bigger theoretical level it breaks down the distinction between social learning of design targets and individual learning of the skills needed to achieve them.
-However, it should be noted that it is not our intention to construct a false binary framework and put these two factors as disconnected and opposite concepts.
-To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
+Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of convergence on shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers.
+At a larger theoretical level it questions the distinction between social learning of design targets vs.
+individual learning of the skills needed to achieve them.
+To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
 In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of bifaces as well as their implications for the biological and cultural evolution of the hominin lineages.

### 2022-08-31 manuscript/manuscript.Rmd
updated the draft, figures, and readme file.

@@
 abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.
-Among many factors identified to contribute to the considerable variation of biface morphology, skill level and mental templates have been frequently cited.
-Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for minimally 0 hours (controlled group) and maximally 90 hours in Late Acheulean style handaxe production and three expert knappers.
+Archaeologists cite many factors that contribute to the considerable variation of biface morphology, including knapper skill levels and mental templates.
+Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
 We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
-Through the principal component analysis of morphometric data derived from images, our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
-Our study demonstrated the potential of experimental archaeology and digital photographs in revealing new insights from old archaeological assemblages.
+Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of biface morphology.
+The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2).
+Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment.
+We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
 \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
@@
 The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019].
-Despite the recurrent narrative emphasizing the homogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021].
-From this extensive list, skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
+Despite the recurrent narrative emphasizing the homogeneity and longevity of biface assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
+More specifically, a complex suite of interconnecting factors have been identified to contribute to biface morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
+From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of biface studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
 Of particular attention here are the experimental works conducted by Stout et al.
@@
 [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage.
-Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage through comparing it with a unique experimental collection.
+Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made bifaces.
 
@@
 For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes.
-To avoid the historical baggage associated with this controversial term, some researchers have also developed alternative frameworks such as "design imperatives" purely derived from ergonomic principles, which refers to a set of minimum features shared by all handaxes including glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
+To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" purely derived from ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
 
-Until very recently, several researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of biface morphology.
+Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of biface morphology.
 Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
@@
 [-@mcnabb2004], which tries to provide a more balanced perspective incorporating both individual agency and social learning.
-As for the critique towards confounding factors explaining morphological variability, raw material is often treated as a crucial variable to be controlled at the very beginning of a research design focusing on mental templates.
+As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates.
 This is best exemplified by an experimental study of García-Medrano et al.
-[-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in archaeological context in composition, size, and shape.
+[-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape.
 In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources.
 Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains.
-In short, when exercised with proper caution, the concept of mental template still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness and elongation, among other things.
+In short, when exercised with proper caution, the concept of mental templates still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
 
@@
 Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers.
-Since the novice biface collection is generated from a 90-hour skill acquisition experiment, we also have the precious opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
-As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental template be efficiently detected from biface morphometric data?
+We generated the novice biface collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
+As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from biface morphometric data?
 2) How does training affect novices' performance in these two aspects?
@@
 
-The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposit [@roberts1998; @pope2020].
-This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominins' subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
+The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020].
+This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
 In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high dexerity reflected in their manufacture.
 As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019].
-To identify the morphological manifestation of knappers' dexterity in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
+To identify the morphological manifestation of knappers' dexterity in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
 The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].
@@
 
-```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the biface measurement protocal using Image J (after Pargeter et al. 2019: Figure 5)."}
+```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the biface measurement protocol using Image J (after Pargeter et al. 2019: Figure 5)."}
 knitr::include_graphics(here::here("figure/Fig1.png"))
@@
 
-Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable explained variance ratio to avoid overfitting.
+Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting.
 Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness.
@@
 
-```{r Screeplot, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
-knitr::include_graphics(here::here("figure/Scree plot.png"))
+```{r Screeplot, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
+knitr::include_graphics(here::here("figure/Fig2.png"))
 ```
@@
 library(tidyverse)
-df1<-read.csv(here::here("data/variable_loadings_pc12.csv"), fileEncoding = 'UTF-8-BOM')
+df1<-read.csv(here::here("data/Table1.csv"), fileEncoding = 'UTF-8-BOM')
 kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components") %>%
@@
 ```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of bifaces from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
-knitr::include_graphics(here::here("figure/General PCA1.png"))
+knitr::include_graphics(here::here("figure/Fig3.png"))
 ```
@@
 ```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132)."}
-knitr::include_graphics(here::here("figure/PC correlation.png"))
+knitr::include_graphics(here::here("figure/Fig4.png"))
 ```
@@
 As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot.
-Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is **not** statistically significant ($t=0.333$, $p>0.05$).
-An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also **not** statistically significant ($t=-0.818$, $p>0.05$).
+Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$).
+An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).
 
-```{r PCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A between-group comparison of PC1 values."}
-knitr::include_graphics(here::here("figure/PC1 comparison.png"))
+```{r PCA1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A between-group comparison of PC1 values."}
+knitr::include_graphics(here::here("figure/Fig5.png"))
 ```
 
-```{r PCA2, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A between-group comparison of PC2 values."}
-knitr::include_graphics(here::here("figure/PC2 comparison.png"))
+```{r PCA2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A between-group comparison of PC2 values."}
+knitr::include_graphics(here::here("figure/Fig6.png"))
 ```
@@
 
-```{r comparison, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the process"}
-knitr::include_graphics(here::here("figure/CAPTURE.PNG"))
+```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the process"}
+knitr::include_graphics(here::here("figure/Fig7.jpg"))
 ```
 
-```{r deltaweight, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A comparison of the delta weight between the pre-training, early training, late training, and the expert group."}
-knitr::include_graphics(here::here("figure/delta weight comparison3.png"))
+```{r deltaweight, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A comparison of the delta weight between the pre-training, early training, late training, and the expert group."}
+knitr::include_graphics(here::here("figure/Fig8.png"))
 ```
@@
 Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions.
-In this paper, the main source of archaeological data is a collection of photos produced and curated by one of our co-authors (R.
-Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original artifacts.
+In this paper, the main source of archaeological data is a collection of photos produced and curated more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original artifacts.
 Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions.

### 2022-09-01 manuscript/manuscript.Rmd
updated the readme file.

@@
 ---
-title: "Detecting skill level and mental templates in Late Acheulean biface morphology: Archaeological and experimental insights"
+title: "Detecting skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
 author: 
@@
 - Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
-abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.
-Archaeologists cite many factors that contribute to the considerable variation of biface morphology, including knapper skill levels and mental templates.
+abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.
+Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates.
 Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
 We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
-Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of biface morphology.
+Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology.
 The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2).
@@
 We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
-\\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
+\\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
 fontsize: 11pt
@@
 
-The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019].
-Despite the recurrent narrative emphasizing the homogeneity and longevity of biface assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors have been identified to contribute to biface morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
-From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of biface studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
+The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019].
+Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
+More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
+From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
 Of particular attention here are the experimental works conducted by Stout et al.
@@
 [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage.
-Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made bifaces.
+Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.
 
@@
 This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework.
-Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applies it in the typological analysis of bifaces at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
+Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applies it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
 Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986].
@@
 
-Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of biface morphology.
+Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of handaxe morphology.
 Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
@@
 Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains.
-In short, when exercised with proper caution, the concept of mental templates still has its value in our study of biface morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
+In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
 
@@
 When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
-In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of biface technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
+In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
 These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].
 
-Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers.
-We generated the novice biface collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
-As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from biface morphometric data?
+Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological handaxe assemblage known for its remarkable dexterity, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
+We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
+As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data?
 2) How does training affect novices' performance in these two aspects?
@@
 
-## Boxgrove biface collection
+## Boxgrove handaxe collection
 
@@
 
-## Experimental biface collection
+## Experimental handaxe collection
 
-The biface experimental replicas used in this study comprised two sub-collection.
-The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014].
+The handaxe experimental replicas used in this study comprised two sub-collection.
+The first sub-collection includes 10 handaxes knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014].
 These handaxes were made for previous research projects, which similarly aimed to approximate 'Late Acheulean' handaxes explicitly comparable to the Boxgrove assemblage [@faisal2010; @stout2011a; @stout2014].
-The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total.
+The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 handaxes in total.
 Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given.
@@
 
-To better understand the morphological variation of Boxgrove biface collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples [@iovita2011], which include both the front and lateral views of a given specimen.
+To better understand the morphological variation of Boxgrove handaxe collection, we adopted a standardized analytical procedure to extract the morphometric information from 752 photos of the studied samples [@iovita2011], which include both the front and lateral views of a given specimen.
 First, we used Adobe Photoshop to conduct a batch transformation of the samples' pixel scale into a real-world measurement scale based on the fixed photographic setting.
 This is then followed by the batch conversion of color photographs to a black-and-white binary format.
-Subsequently, we cropped the silhouettes of bifaces one by one using the Quick Selection Tool in Adobe Photoshop.
+Subsequently, we cropped the silhouettes of handaxes one by one using the Quick Selection Tool in Adobe Photoshop.
 The metric measurements were conducted in ImageJ [@rueden2017], where we employed a custom ImageJ script [@pargeter2019] to measure the maximum length, width, and thickness of a given silhouette.
-The width and thickness measurements are taken at 10% increments of length starting at the tip of each biface (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements).
+The width and thickness measurements are taken at 10% increments of length starting at the tip of each handaxe (**Figure** \@ref(fig:ImageJ)), which eventually leads to 19 morphometric variables in total (1 length measurement, 9 width measurements, and 9 thickness measurements).
 Finally, we calculated the geometric means of all 19 linear measurements to create a scale-free data set that preserves the individual morphological variation at the same time [@lycett2006].
@@
 Notably, Shipton and Clarkson [-@shipton2015] previously found that reduction intensity does not have a strong impact on the shape of handaxes.
-The same procedure was also applied to the morphometric analyses of the experimental biface collection, which was partially published in Pargeter et al.
+The same procedure was also applied to the morphometric analyses of the experimental handaxe collection, which was partially published in Pargeter et al.
 [-@pargeter2019].
 
-```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the biface measurement protocol using Image J (after Pargeter et al. 2019: Figure 5)."}
+```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the handaxe measurement protocol using Image J (after Pargeter et al. 2019: Figure 5)."}
 knitr::include_graphics(here::here("figure/Fig1.png"))
@@
 Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017].
-To detect the effect of training on novices' performance as compared with archaeological samples and biface made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance.
+To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance.
 This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github (<https://github.com/Raylc/PaST-pilot>).
@@
 It is positively correlated with all thickness measurements while negatively correlated with all other measurements.
-A higher PC1 value thus indicates a thicker biface, and vice versa.
+A higher PC1 value thus indicates a thicker handaxe, and vice versa.
 The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness.
-As PC2 increases, a biface will be relatively longer and more convergent from the broad base to the tip..
+As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip..
 Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.
@@
 
-A closer look at the principal component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) yields the clustering of different groups of bifaces.
-The majority of Boxgrove bifaces occupy an area featuring negative values of both PC1 and PC2.
+A closer look at the principal component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) yields the clustering of different groups of handaxes.
+The majority of Boxgrove handaxes occupy an area featuring negative values of both PC1 and PC2.
 The expert group is similar to the Boxgrove group in PC1, while the former has a relatively higher PC2 value than the latter on average.
-The group of novice displays the highest level of variability, however, it is rather pronounced that most bifaces made by novices have a positive PC1 value that is different from both the groups of Boxgrove and experts.
+The group of novice displays the highest level of variability, however, it is rather pronounced that most handaxes made by novices have a positive PC1 value that is different from both the groups of Boxgrove and experts.
 
-```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of bifaces from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
+```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of handaxes from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
 knitr::include_graphics(here::here("figure/Fig3.png"))
@@
 
-We extracted the PC1 and PC2 values of individual bifaces and compared them between different groups.
+We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups.
 More specifically, the novice group was divided into three sub-groups based on their training stages as specified in the method section.
@@
 
-Our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
+Our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
 Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
-In accordance with the existing literature on biface knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers.
-Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the biface into several pieces, serving the purpose of achieving the desired convexity and/or volume.
+In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers.
+Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume.
 This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013].
-Experimental studies have also shown that the thinning stage of biface produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394].
+Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394].
 This also reflects the majority of samples in both our expert and novice experimental assemblages.
@@
 [-@garcía-medrano2019].
-This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in biface analysis[-@hutchence2021].
+This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis[-@hutchence2021].
 The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.
 Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions.
-Unfortunately, the biface data from the instructor (N.
+Unfortunately, the handaxe data from the instructor (N.
 Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley).
@@
 To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
-In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of bifaces as well as their implications for the biological and cultural evolution of the hominin lineages.
+In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes as well as their implications for the biological and cultural evolution of the hominin lineages.
 

### 2022-10-17 manuscript/manuscript.Rmd
deleted photos and corrected the repository web link.

@@
 To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance.
-This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github (<https://github.com/Raylc/PaST-pilot>).
+This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).
 

### 2022-11-06 manuscript/manuscript.Rmd
minor revisions.

@@
 ---
-title: "Detecting skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
+title: "Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
 author: 
@@
 Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
+More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillinger2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
 From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
@@
 
+We would like to thank Thomas Jennings and three other anonymous reviewers for their insightful feedback on an earlier draft of this manuscript.
 This work was supported by funding from the National Science Foundation of the USA (grants SMA-1328567 & DRL-1631563), the John Templeton Foundation (grant 47994), and the Emory University Research Council.

### 2022-11-07 manuscript/manuscript.Rmd
minor revisions.

@@
 This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework.
-Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applies it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
+Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
 Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986].
 For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes.
-To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" purely derived from ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
+To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
 
@@
 
-Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template and the interaction between the two through a comparative study of an archaeological handaxe assemblage known for its remarkable dexterity, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
+Drawing on these two lines of literature, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
 We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
@@
 This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
-In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high dexerity reflected in their manufacture.
+In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture.
 As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019].
-To identify the morphological manifestation of knappers' dexterity in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
+To identify the morphological manifestation of knappers' skill level in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
 The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].
@@
 
-In this experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas from our expert sample.
+In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas from our expert sample.
 Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage.
@@
 
-In addition, visual inspection of the principle component scatter plot (Fig.
-) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups.
+In addition, visual inspection of the principle component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups.
 To test this, we conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)).

### 2022-11-18 manuscript/manuscript.Rmd
addressed several minor comments from Reviewer #2.

@@
 Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillinger2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
+More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillinger2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
 From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
@@
 Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986].
-For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes.
+For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], and b) ignorance of other competing factors such as raw material constraints [@white1995].
 To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
 
-Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of handaxe morphology.
+Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology.
 Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
@@
 This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al.
-[-@mcnabb2004], which tries to provide a more balanced perspective incorporating both individual agency and social learning.
+[-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning.
 As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates.
@@
 [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape.
-In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources.
-Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains.
 In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
@@
 
+As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest.
 Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017].
-To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance.
-This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).
+To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test.
+Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test.
+The sample size of each compared group can be as low as 6 [@games1976; @sauder2019].
+This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).
 
@@
 The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness.
-As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip..
+As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip.
 Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.
@@
 df1<-read.csv(here::here("data/Table1.csv"), fileEncoding = 'UTF-8-BOM')
-kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components") %>%
+df1<- df1 %>% 
+  mutate_if(is.numeric, round, digits = 4) %>% 
+  rename(Variables = X)
+kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components.") %>%
   kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
@@
 
-We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups.
-More specifically, the novice group was divided into three sub-groups based on their training stages as specified in the method section.
+We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section.
 As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot.

### 2022-11-20 manuscript/manuscript.Rmd
addressed some minor comments from Reviewer 2.

@@
 
-Our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
+Our study suggests that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
 Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
@@
 This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013].
-Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394].
-This also reflects the majority of samples in both our expert and novice experimental assemblages.
-In the skill acquisition experiments, novice knappers have been explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment.
-On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], corresponding to the cases of replicas produced by Bruce Bradley.
+Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394], although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014].
+It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages.
+In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment.
+On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.
 

### 2022-11-21 manuscript/manuscript.Rmd
changed the caption of Table 1.

@@
   rename(Variables = X)
-kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components.") %>%
+kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components.
+PC1 (Dim.1) is postively correlated with all thickness-related variables and negatively correlated with all width-related variables and the maximum length.
+PC2 (Dim.2) is positively with bottom width and thickness variables as well as the miximum length and negatively correlated with width and thickness variables of the tip area.") %>%
   kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%

### 2022-11-23 manuscript/manuscript.Rmd
minor revision based on R2.

@@
 
-The pre-training group is similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
+The pre-training group is unexpectedly similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
 If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged.

### 2022-11-23 manuscript/manuscript.Rmd
finished addressing all minor comments by R2.

@@
 As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data?
+Accordingly, we hypothesize that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences.
 2) How does training affect novices' performance in these two aspects?
+Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template.
 
@@
 An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).
+These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (skill level) and PC2 (mental template).
+However, the effects of training across different assessment periods on both dimensions are not significant.
 
@@
 [-@garcía-medrano2019].
-This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis[-@hutchence2021].
+This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis [-@hutchence2021].
 The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.
@@
 post-training) in both PC1 (skill level) and PC2 (mental template).
-Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather inconsipicous.
+Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant.
 This finding corroborates what has been suggested in Pargeter et al.
@@
 This follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue.
-It is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
-Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice[@milks2019].
+Corroborating Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
+Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].
 

### 2022-11-23 code/Boxgrove_Iovita.R

@@ -317,9 +317,11 @@ p1<-ggstatsplot::ggbetweenstats(
   x     = assessment_stage,
   y     = PC1,
   ggsignif.args = list(textsize = 1.5, tip_length = 0.01),
+  results.subtitle = FALSE,
   title = "A between-group comparison of PC1 values"
 )
-ggplot2::ggsave("PC1 comparison.png", path="figure", dpi = 600)
+ggplot2::ggsave("PC1 comparison.png", path="figure", width = 20,
+                height = 10, units = "cm", dpi = 600)
 ggstatsplot::extract_stats(p1)
 
 p2<-ggstatsplot::ggbetweenstats(
@@ -327,9 +329,11 @@ p2<-ggstatsplot::ggbetweenstats(
   x     = assessment_stage,
   y     = PC2,
   ggsignif.args = list(textsize = 1.5, tip_length = 0.01),
+  results.subtitle = FALSE,
   title = "A between-group comparison of PC2 values"
 )
-ggplot2::ggsave("PC2 comparison.png", path="figure", dpi = 600)
+ggplot2::ggsave("PC2 comparison.png", path="figure", width = 20,
+                height = 10, units = "cm", dpi = 600)
 ggstatsplot::extract_stats(p2)
 
 
@@ -369,7 +373,7 @@ total_grouped <-total2 %>%
                                  ifelse(Assessment %in% c("2","3","4","5"),"Early training",
                                         ifelse(Assessment %in% c("6","7","8","9"),"Late training", 
                                           ifelse(Assessment == "10","Expert","Boxgrove")))))
-write.csv(total_grouped,"data/Experiment/Weight data.csv", row.names = FALSE)
+# write.csv(total_grouped,"data/Experiment/Weight data.csv", row.names = FALSE)
 
 #reorder the dataframe based on training stages for plotting.
 corder <- c("Expert", "Late training", "Early training", "Pre-training")
@@ -384,25 +388,10 @@ p <- ggstatsplot::ggbetweenstats(
   x     = Assessment_stage,
   y     = delta_Weight,
   ggsignif.args = list(textsize = 1.5, tip_length = 0.01),
+  results.subtitle = FALSE,
   title = "A between-group comparison of delta weight")
-ggplot2::ggsave("delta weight comparison3.png", path="figure", dpi = 600)
+ggplot2::ggsave("delta weight comparison3.png", path="figure", width = 20,
[…]

### 2022-11-24 manuscript/manuscript.Rmd
finished addressing R1's comments.

@@
 Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillinger2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
+More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillingerConsideringRoleTime2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
 From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
@@
 Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986].
-For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], and b) ignorance of other competing factors such as raw material constraints [@white1995].
+For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016].
 To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
+The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of internal representation, where handaxes can emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]."
 
@@
 [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape.
-In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
+Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm.
+However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
 
@@
 Consequently, it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed.
-It is possible that such constraints help to explain the convergence of our novice knappers on similar shapes to those prefered by modern expert knappers, however, this clearly does not explain the design target at Boxgrove.
+It is possible that such constraints help to explain the convergence of our novice knappers on similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove.
 Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers.
@@
 Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant.
-This finding corroborates what has been suggested in Pargeter et al.
+This finding provides a parallel line of evidence that corroborates what has been suggested in Pargeter et al.
 [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures.
-This follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue.
-Corroborating Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
-Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].
+Methodologically speaking, this study also demonstrated that the pattern revealed by the multivariate analysis of morphometric data can nicely match with the expert knapper's 5 point grading scale of novices' knapping performances that takes multiple factors into consideration, including outcome, perceptual motor execution, and strategic understanding [See Table 2 of -@pargeter2019 for more details].
+
+Moreover, this follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue.
+As previously shown in Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** @ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
+Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; 2) the present-day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019]; and/or 3) modern knappers' skill level was affected by time constraints when they were requested to produce the reference collections [@schillingerConsideringRoleTime2014; @lewis2022].
 
@@
 
-Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions.
+Although we are not the first research team to use secondary archaeological data [e.g., @keyHandaxeShapeVariation2019], we would still like to highlight here that this research project further exemplifys the potential of reusing old archaeological data in digital format to address novel research questions.
 In this paper, the main source of archaeological data is a collection of photos produced and curated more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original artifacts.
@@
 To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
-In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes as well as their implications for the biological and cultural evolution of the hominin lineages.
+In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.
 

### 2022-11-26 manuscript/manuscript.Rmd
minor revisions.

@@
 Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillingerConsideringRoleTime2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
+More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
 From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
@@
 This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of the mental template in that they both refute the idea that technology is simply an internal program expressed by the mind and they prefer a dynamic approach emphasizing the interaction between perception and action.
-The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners.
+The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners.
 It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].
@@
 In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture.
-As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019].
+As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019].
 To identify the morphological manifestation of knappers' skill level in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
@@
 The expert group is similar to the Boxgrove group in PC1, while the former has a relatively higher PC2 value than the latter on average.
-The group of novice displays the highest level of variability, however, it is rather pronounced that most handaxes made by novices have a positive PC1 value that is different from both the groups of Boxgrove and experts.
+The group of novice displays the highest ranges in both PC1 and PC2 values according to the scatter plot, however, it is rather pronounced that most handaxes made by novices have a positive PC1 value that is different from both the groups of Boxgrove and experts.
 

### 2022-11-27 manuscript/manuscript.Rmd
changed the caption of figure 4.

@@
 
-```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132)."}
+```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132).
+The upper left area in each individual plot display statistical reporting from a frequentist perspective, including the student-t test statistics, p-value, Pearson correlation coefficient, confidence interval, and sample size.
+The lower right area in each individual plot display statistical reporting from a Bayesian perspective, including the natural logarithm of Bayes factor, posterior type and estimate, credible interval, and prior type and value.
+"}
 knitr::include_graphics(here::here("figure/Fig4.png"))

### 2022-11-28 manuscript/manuscript.Rmd
addressed some major comments of R2.

@@
 
+We use the statistical programming language R 4.1.1 [@rcoreteam2021] to conduct statistical analyses and data visualization in this study, particularly the R packages "FactoMineR" [@lê2008] and "ggstatsplot" [@patil2021].
 As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest.
@@
 ```{r PCcorrelation, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A scatter plot showing the correlation between PC1 and PC2 respectively in the groups of Boxgrove (left, n=326), expert (middle, n=10), and novice (right, n=132).
-The upper left area in each individual plot display statistical reporting from a frequentist perspective, including the student-t test statistics, p-value, Pearson correlation coefficient, confidence interval, and sample size.
-The lower right area in each individual plot display statistical reporting from a Bayesian perspective, including the natural logarithm of Bayes factor, posterior type and estimate, credible interval, and prior type and value.
+The upper left area in each individual plot displays statistical reporting from a frequentist perspective, including the student-t test statistics, p-value, Pearson correlation coefficient, confidence interval, and sample size.
+The lower right area in each individual plot displays statistical reporting from a Bayesian perspective, including the natural logarithm of Bayes factor, posterior type and estimate, credible interval, and prior type and value.
 "}
@@
 Our study suggests that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
+Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and confirmed the previous finding [-@pargeter2019] that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
 In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers.
@@
 The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form.
-Our results (Fig.) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value).
+Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes.
 In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface.
@@
 
-```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the process"}
+```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the knapping process."}
 knitr::include_graphics(here::here("figure/Fig7.jpg"))

### 2022-11-28 manuscript/manuscript.Rmd
finished addressing all R2's comments except research question rephrasing.

@@
 The sample size of each compared group can be as low as 6 [@games1976; @sauder2019].
+Lastly, we compare the delta weight, as defined by the difference between initial nodule weight and end product weight, between these groups to understand the effect of training on reduction intensity.
 This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).
@@
 However, the effects of training across different assessment periods on both dimensions are not significant.
+Regarding the delta weight of different groups, our analysis (**Figure** \@ref(fig:weight)) suggests that there is a significant difference between the pre-training group and Late training group, while all other pairwise group comparison results are insignificant.
+It can also be inferred that the expert group display a higher variability in terms of delta weight compared with novices.
 
@@
 knitr::include_graphics(here::here("figure/Fig6.png"))
+```
+
+```{r weight, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A comparison of the delta weight between the pre-training, early training, late training, and the expert group."}
+knitr::include_graphics(here::here("figure/Fig8.png"))
 ```
@@
 Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes.
-In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface.
-Consequently, it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed.
-It is possible that such constraints help to explain the convergence of our novice knappers on similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove.
+In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90].
+As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed.
+It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove.
 Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers.
-This also provides an alternative explanation to the social transmission of form for the experimental convergence of on pointed forms.
+This also provides an alternative explanation to the social transmission of form for the experimental convergence on pointed forms.
 In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.
@@
 Moreover, this follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue.
-As previously shown in Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** @ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
+As previously shown in Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
 Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; 2) the present-day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019]; and/or 3) modern knappers' skill level was affected by time constraints when they were requested to produce the reference collections [@schillingerConsideringRoleTime2014; @lewis2022].
@@
 The pre-training group is unexpectedly similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
-If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged.
-This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:deltaweight)).
+If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005].
+This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)).
 It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape.
 Experts generally try to achieve handaxe forms while removing as little mass as possible (i.e.
-making as big a handaxe as possible from the nodule).
+making as big a handaxe as possible from the nodule) [@whittaker1994].
 On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360].
 These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>).
-As such, behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.
+As such, we infer that behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.
 
@@
 knitr::include_graphics(here::here("figure/Fig7.jpg"))
-```
-
-```{r deltaweight, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A comparison of the delta weight between the pre-training, early training, late training, and the expert group."}
-knitr::include_graphics(here::here("figure/Fig8.png"))
 ```
@@
 
-Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of convergence on shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers.
+Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers.
 At a larger theoretical level it questions the distinction between social learning of design targets vs.

### 2022-11-28 code/Boxgrove-Exp-code.R

@@ -1,333 +1,0 @@
-############################## 
-# Add functions
-##############################
-library(tidyverse)
-library(reshape2)
-library(psych)
-library(factoextra)
-library(ggfortify)
-require(FactoMineR)
-library(data.table)
-library(sjmisc)
-library(sjPlot)
-require(MASS)
-require(cowplot)
-library(sjstats)
-library(pwr)
-############################## 
-# Set working directory and source files
-##############################
-
-# Set to source of shape and size data txt files
-getwd()
-
-############################## 
-# Add necessary datasets
-##############################
-
-# Size/shape data
-plan_filenames_162<-list.files("data/Boxgrove_Apel/Measurements/plan_scale_162", pattern="*.txt")
-profile_filenames_162<-list.files("data/Boxgrove_Apel/Measurements/profile_scale_162", pattern="*.txt")
-
-# Size/area data
-area_filenames<-list.files("data/Boxgrove_Apel/Measurements/Area", pattern = "*.csv")
-
-#experimental handaxe data
-experimental_data<-read.csv("data/Experiment/experimental_handaxes.csv")
-
-experimental_data<-experimental_data %>%
-  rename(individual = knapper) %>%
-  mutate(group=ifelse(individual %in% c(1:21),"novice",
-                      ifelse(individual %in% c(22:37),"control","expert")))
-
-##############################
-# Open Read function
-# Calls txt and csv files into R
-##############################
-
-# open txt files function
-open_read<-function(filename, foldername){
-  open_file<-read.delim(paste(foldername,"/",filename,sep=""),header=F,sep='_')
-  colnames(open_file)<-c("variable","measurement_point","measurement")
-  open_file$individual<-gsub(".txt","",filename)
-  return(open_file)
-}
-
-##############################
-# Merge shape data function
-# Merges size data for shape analysis
-##############################
-# Merge plans and profiles of silhouette with 162-cm-scale
-merge_shape_data_function_162<-function(plan_names,profile_names,plan_pathway,profile_pathway) {
-  
-  # combine txt files for each recording set
-  handaxe_plan_measurements<-do.call(rbind,lapply(plan_names,open_read,plan_pathway))
[…]

### 2022-11-29 manuscript/manuscript_word.Rmd
finished addressing all comments except restructring research questions and created a word version.

--- 2022-11-28 manuscript/manuscript.Rmd
@@
 ---
-title: "Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
+title: 'Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights'
 author:
@@
 We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
-\\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
-fontsize: 11pt
-output: bookdown::pdf_document2
-header-includes:
-   - \usepackage{textcomp}
-   - \usepackage{utopia}
-   - \usepackage{lineno}
-   - \linenumbers
-   - \usepackage{float}
-   - \floatplacement{figure}{H}
-   - \usepackage{caption}
-   - \captionsetup[figure]{font=scriptsize}
-   - \captionsetup[table]{font=scriptsize}
+Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
+output: word_document
 linestretch: 1.5
@@
 csl: apa.csl
+fontsize: 11pt
+always_allow_html: true
 ---
@@
 Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
-More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021].
+More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017].
 From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
@@
 As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data?
-Accordingly, we hypothesize that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences.
+Accordingly, we hypothesize that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is the common target.
 2) How does training affect novices' performance in these two aspects?
@@
 It is positively correlated with all thickness measurements while negatively correlated with all other measurements.
-A higher PC1 value thus indicates a thicker handaxe, and vice versa.
+A higher PC1 value thus indicates a handaxe with thicker shape relative to width and length, and vice versa.
 The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness.
@@
   rename(Variables = X)
+df1[1, 1] <- "width_90%"
+df1[2, 1] <- "width_80%"
+df1[3, 1] <- "width_70%"
+df1[4, 1] <- "width_60%"
+df1[5, 1] <- "width_50%"
+df1[6, 1] <- "width_40%"
+df1[7, 1] <- "width_30%"
+df1[8, 1] <- "width_20%"
+df1[9, 1] <- "width_10%"
+df1[10, 1] <- "thickness_90%"
+df1[11, 1] <- "thickness_80%"
+df1[12, 1] <- "thickness_70%"
+df1[13, 1] <- "thickness_60%"
+df1[14, 1] <- "thickness_50%"
+df1[15, 1] <- "thickness_40%"
+df1[16, 1] <- "thickness_30%"
+df1[17, 1] <- "thickness_20%"
+df1[18, 1] <- "thickness_10%"
 kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components.
@@
 [-@garcía-medrano2019].
-This pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis [-@hutchence2021].
+The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe."In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis [-@hutchence2021].
 The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.
@@
 The pre-training group is unexpectedly similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
-If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005].
-This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)).
+If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113].
+This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** @ref(fig:weight)).
 It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape.
-Experts generally try to achieve handaxe forms while removing as little mass as possible (i.e.
-making as big a handaxe as possible from the nodule) [@whittaker1994].
+Achieve handaxe forms while removing as little mass as possible (i.e.
+making as big a handaxe as possible from the nodule) generally requires a higher skill level due to the reductive or subtractive nature of stone knapping, where correcting an error or any thinning procedure always requires the removal of raw material and thereby reducing the size of a given handaxe [@schillingerCopyingErrorCultural2014a: 130; @deetz1967: 48-49].
 On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by knappers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360].
@@
 To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
+Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizibility to archaeological data).
+In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022].
 In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.

### 2022-12-01 manuscript/manuscript.Rmd
added the handaxe photo as fig1.

@@
 \\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
+date: "`r Sys.Date()`"
 fontsize: 11pt
@@
 This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
-In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture.
+In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
 As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019].
@@
 
+```{r photos, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A selection of Boxgrove handaxes and modern replicas produced by experts and novices."}
+knitr::include_graphics(here::here("figure/Fig1.jpg"))
+```
+
 ## Experimental handaxe collection
 
-The handaxe experimental replicas used in this study comprised two sub-collection.
+The handaxe experimental replicas used in this study comprised two sub-collection (**Figure** \@ref(fig:photos)).
 The first sub-collection includes 10 handaxes knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014].
@@
 
-```{r ImageJ, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A visual demonstration of the handaxe measurement protocol using Image J (after Pargeter et al. 2019: Figure 5)."}
-knitr::include_graphics(here::here("figure/Fig1.png"))
+```{r ImageJ, echo=FALSE, out.width = '80%', out.height = '50%', fig.align='center', fig.cap="A visual demonstration of the handaxe measurement protocol using Image J (after Pargeter et al.
+2019: Figure 5)."}
+knitr::include_graphics(here::here("figure/Fig2.png"))
 ```
@@
 ```{r Screeplot, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A scree plot showing the percentage of explained variances of the first 10 principal components."}
-knitr::include_graphics(here::here("figure/Fig2.png"))
+knitr::include_graphics(here::here("figure/Fig3.png"))
 ```
@@
 ```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of handaxes from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
-knitr::include_graphics(here::here("figure/Fig3.png"))
+knitr::include_graphics(here::here("figure/Fig4.png"))
 ```
@@
 "}
-knitr::include_graphics(here::here("figure/Fig4.png"))
+knitr::include_graphics(here::here("figure/Fig5.png"))
 ```
@@
 ```{r PCA1, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A between-group comparison of PC1 values."}
-knitr::include_graphics(here::here("figure/Fig5.png"))
+knitr::include_graphics(here::here("figure/Fig6.png"))
 ```
@@
 ```{r PCA2, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A between-group comparison of PC2 values."}
-knitr::include_graphics(here::here("figure/Fig6.png"))
+knitr::include_graphics(here::here("figure/Fig7.png"))
 ```
@@
 ```{r comparison, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="Core 63 before (left) and after knapping(right), showing the minimal morphological change during the knapping process."}
-knitr::include_graphics(here::here("figure/Fig7.jpg"))
+knitr::include_graphics(here::here("figure/Fig9.jpg"))
 ```
@@
 To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
+Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015].
+This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data).
+In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022].
 In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.

### 2022-12-01 manuscript/manuscript_word.Rmd
added the handaxe photo as fig1.

@@
 Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
-output: word_document
+output:
+  word_document:
+    reference_docx: "manuscript word.docx"
 linestretch: 1.5
@@
 To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
-Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizibility to archaeological data).
+Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015].
+This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data).
 In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022].

### 2022-12-08 manuscript/manuscript.Rmd
affiliation changes.

@@
 - Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
-- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
+- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; ParkRock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
 abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.

### 2022-12-10 manuscript/manuscript.Rmd
minor correction.

--- 2022-12-01 manuscript/manuscript.Rmd
@@
 - Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
-- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
+- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
 abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.
@@
 Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment.
-We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
-\\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
+We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
 date: "`r Sys.Date()`"

### 2022-12-25 manuscript/manuscript.Rmd
minor revision including title change.

--- 2022-12-01 manuscript/manuscript.Rmd
@@
 ---
-title: "Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
+title: "Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
 author: 
@@
 Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment.
-We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
-\\par \\textbf{Keywords:} Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
+We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural reproduction"
 date: "`r Sys.Date()`"
@@
 Of particular attention here are the experimental works conducted by Stout et al.
-[-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al.
+[-@stout2014] focusing on inferring knapping skill level and Garcia-Medrano et al.
 [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage.
@@
 To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
-The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of internal representation, where handaxes can emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]."
+The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]."
 
-Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology.
-Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
+Recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology.
+Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations.
+From this perspective, social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
 They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms.
@@
 When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
-In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
+In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019; @whittaker2004: 180-182] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
 These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].
@@
 Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting.
-Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness.
+Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures relative cross-sectional thickness ("refinement").
 It is positively correlated with all thickness measurements while negatively correlated with all other measurements.
-A higher PC1 value thus indicates a handaxe with thicker shape relative to width and length, and vice versa.
+A higher PC1 value thus indicates a handaxe that is thicker relative to width and length, and vice versa.
 The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness.
 As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip.
-Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.
+Thus, PC1 corresponds to cross-sectional thinning and PC2 to a narrowing of the tip relative to length and base dimensions.
 
@@
 An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).
-These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (skill level) and PC2 (mental template).
+These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (thinning) and PC2 (pointedness).
 However, the effects of training across different assessment periods on both dimensions are not significant.
@@
 In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90].
-As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed.
+As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182].
 It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove.

### 2022-12-28 manuscript/manuscript.Rmd
added a few more recent references.

@@
 [-@garcía-medrano2019].
-The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe."In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis [-@hutchence2021].
+The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe."In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide an mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020].
 The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.
@@
 If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113].
-This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** @ref(fig:weight)).
+This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)).
 It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape.
@@
 
-**Cheng Liu**: Data curation, Formal analysis, Investigation, Methodology, Visualization, Writing -- original draft, Writing -- review & editing.
+**Cheng Liu**: Conceptualization, Data curation, Formal analysis, Investigation, Methodology, Visualization, Writing -- original draft, Writing -- review & editing.
 **Nada Khreisheh**: Investigation, Writing -- review & editing.

### 2022-12-31 manuscript/manuscript.Rmd
minor changes in keywords and references.

@@
 Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment.
-We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural reproduction"
+We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
 date: "`r Sys.Date()`"
@@
 This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
-In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
+In addition to the presence of one of the earliest hominin fossil [tentatively assigned to *Homo heidelbergensis*, @hillson2010; @lockey2022] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
 As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019].

### 2022-12-31 manuscript/manuscript.Rmd
added a few more references.

@@
 The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020].
-This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
-In addition to the presence of one of the earliest hominin fossil [tentatively assigned to *Homo heidelbergensis*, @hillson2010; @lockey2022] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
+This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010; @preece2022].
+In addition to the presence of one of the earliest hominin fossil [tentatively assigned to *Homo heidelbergensis*, @hillson2010; @lockey2022; @roberts1994] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
 As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019].

### 2023-01-07 manuscript/manuscript.Rmd
added a few more references.

@@
 The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019].
-Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011].
+Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011; @moncel2018a; @moncel2018b; @moncel2018c].
 More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017].

### 2023-01-20 manuscript/manuscript.Rmd
finished another round of revision.

@@
 Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates.
-Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
+Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
 We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
@@
 The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment.
-We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
+Moreover, we also evaluated the effects of training on the differential cultural reproduction of these two aspects using the data from a 90-hour-long knapping skill acquisition experiment.
+We found that the desired shape of a handaxe can be relatively quickly picked up by novices, while reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
 date: "`r Sys.Date()`"
@@
 [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage.
-Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.
+Our paper incorporates these two perspectives into a broader conceptual framework of cultural reproduction and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.
+
+## Mental template
 
@@
 
+## Skill level
+
 Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
@@
 
-Drawing on these two lines of literature, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
+## Cultural reproduction
+
+The cultural reproduction, or cultural transmission as described in standard cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and skill levels makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records.
+Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed.
+The ignorance of the latter becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a].
+This reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].
+
+Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
 We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
 As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data?
-Accordingly, we hypothesize that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is the common target.
-2) How does training affect novices' performance in these two aspects?
-Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template.
+Accordingly, we predict that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is a common target.
+2) Does training has a differential effect in terms of the reproduction of skill level and mental templates among novices?
+Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby slower than the latter aspect.
+This hypothesis is informed by the previous study of Pargeter et al.
+[-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.
 
@@
 The sample size of each compared group can be as low as 6 [@games1976; @sauder2019].
-Lastly, we compare the delta weight, as defined by the difference between initial nodule weight and end product weight, between these groups to understand the effect of training on reduction intensity.
+Lastly, we compare the delta weight, as defined by the difference between initial nodule weight and end product weight, between these groups to understand the effect of reduction intensity on morphological variation.
 This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).
@@
 Our study suggests that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and confirmed the previous finding [-@pargeter2019] that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
+Moreover, we also evaluated the differential effects of training using the data from a 90-hour long knapping skill acquisition experiment and confirmed the previous finding [-@pargeter2019] that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
 In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers.
@@
 Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis.
+It should be noted that a thin cross section as measured by PC1 could also be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, however, novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
 Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al.
 [-@garcía-medrano2019].
-The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe."In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide an mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020].
+The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe." In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide an mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020].
 The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.
@@
 Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant.
+When the performance of experts is used as a reference point here, we can see that for PC2 no significance difference is detected between early training, late training, and expert group, while for PC1 the expert group is clearly different from the training groups, supporting our hypothesis in terms of the differential cultural reproduction of mental templates and skill level.
 This finding provides a parallel line of evidence that corroborates what has been suggested in Pargeter et al.
@@
 
-Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers.
+Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers, which supports our differential cultural reproduction hypothesis.
 At a larger theoretical level it questions the distinction between social learning of design targets vs.
 individual learning of the skills needed to achieve them.
-To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
 Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015].
@@
 In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022].
-In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.
+In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.
 

### 2023-01-20 manuscript/manuscript_word.Rmd
finished another round of revision.

@@
 The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment.
-We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
-Late Acheulean; Handaxe production; Boxgrove; Experimental archaeology; Skill level; Mental template"
+Moreover, we also evaluated the effects of training on the differential cultural reproduction of these two aspects using the data from a 90-hour-long knapping skill acquisition experiment.
+We found that the desired shape of a handaxe can be relatively quickly picked up by novices, while reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
 output:
@@
 Of particular attention here are the experimental works conducted by Stout et al.
-[-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al.
+[-@stout2014] focusing on inferring knapping skill level and Garcia-Medrano et al.
 [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage.
-Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.
+Our paper incorporates these two perspectives into a broader conceptual framework of cultural reproduction and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.
+
+## Mental template
 
@@
 When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
-In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
+In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019; @whittaker2004: 180-182] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
 These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].
 
-Drawing on these two lines of literature, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
+## Cultural reproduction
+
+The cultural reproduction, or cultural transmission as described in standard cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and skill levels makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records.
+Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed.
+The ignorance of the latter becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a].
+This reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].
+
+Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
 We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
 As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data?
-Accordingly, we hypothesize that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is the common target.
-2) How does training affect novices' performance in these two aspects?
-Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template.
+Accordingly, we predict that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is a common target.
+2) Does training has a differential effect in terms of the reproduction of skill level and mental templates among novices?
+Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby slower than the latter aspect.
+This hypothesis is informed by the previous study of Pargeter et al.
+[-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.
 
@@
 The archaeological site of Boxgrove is located in the former Eartham quarry, Boxgrove, West Sussex, featuring a long sequence of Middle Pleistocene deposits [@roberts1998; @pope2020].
-This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010].
-In addition to the presence of one of the earliest hominin fossil [*Homo heidelbergensis*, @hillson2010] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
+This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010; @preece2022].
+In addition to the presence of one of the earliest hominin fossil [tentatively assigned to *Homo heidelbergensis*, @hillson2010; @lockey2022; @roberts1994] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
 As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019].
@@
 Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis.
+It should be noted that a thin cross section as measured by PC1 could also be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, however, novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
 Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al.
 [-@garcía-medrano2019].
-The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe."In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in handaxe analysis [-@hutchence2021].
+The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe." In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide an mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020].
 The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.

### 2023-01-23 manuscript/manuscript.Rmd
updated the fig4 by inserting image illustration.

@@
 
-```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of handaxes from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132)."}
+```{r GeneralPCA1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="A principal component scatter plot of handaxes from the groups of Boxgrove (red, n=326), expert (green, n=10), and novice (blue, n=132).
+The four images illustrate simplified plan and profile morphology of handaxes displaying extreme PC values (e.g., The leftmost and uppermost handaxes respectively display the highest PC1 and PC2 value, and vice versa)."}
 knitr::include_graphics(here::here("figure/Fig4.png"))

### 2023-01-23 manuscript/manuscript_word.Rmd
updated the fig4 by inserting image illustration.

@@
 ---
-title: 'Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights'
+title: 'Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights'
 author:

### 2023-01-23 code/Boxgrove_Iovita.R

@@ -97,7 +97,7 @@ merge_shape_data_function<-function(plan_names,profile_names,plan_pathway,profil
 boxgrove_measurements<-merge_shape_data_function(plan_filenames,profile_filenames,"data/Boxgrove_Iovita/Measurements/plan","data/Boxgrove_Iovita/Measurements/profile")
 
 # Save the merged data as a csv file
-write.csv(boxgrove_measurements,"data/Boxgrove_Iovita/Measurements/merged.csv", row.names = FALSE)
+#write.csv(boxgrove_measurements,"data/Boxgrove_Iovita/Measurements/merged.csv", row.names = FALSE)
 
 ##############################
 # Open Read size/area function
@@ -183,12 +183,30 @@ bepca <-subset(boxgrove_experiment_data_shapes, select=-c(assessment,group))
 bepca <-prcomp(boxgrove_experiment_data_shapes[,-c(1:3)],  scale = TRUE)
 bepca1 <-prcomp(boxgrove_experiment_data_shapes[,-c(1:3)],  scale = FALSE)
 
-fviz_pca_ind(bepca1, habillage= boxgrove_experiment_data_shapes1$group, # color by groups
+pcaplot<-fviz_pca_ind(bepca1, habillage= boxgrove_experiment_data_shapes1$group, # color by groups
              addEllipses = TRUE, # Concentration ellipses
              ellipse.type = "convex",
              legend.title = "Groups",
              label="none")
 ggsave("General PCA1.png", path="figure", dpi = 600)
+
+fviz_pca_ind(bepca1)
+library(cowplot)
+
+pcaplot1<-ggdraw(pcaplot) + 
+  draw_image("figure/HPC1.PNG",
+             x = .68, y = 0.16, width = 0.12, height = 0.12)
+pcaplot2<-ggdraw(pcaplot1) + 
+  draw_image("figure/LPC1.PNG",
+             x = .03, y = 0.16, width = 0.12, height = 0.12)
+pcaplot3<-ggdraw(pcaplot2) + 
+  draw_image("figure/LPC2.PNG",
+             x = .3, y = 0.05, width = 0.12, height = 0.12)
+pcaplot4<-ggdraw(pcaplot3) + 
+  draw_image("figure/HPC2.PNG",
+             x = .3, y = 0.82, width = 0.12, height = 0.12)
[…]

### 2023-01-28 manuscript/manuscript.Rmd
finished another round of revision.

@@
 ---
-title: "Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
+title: "Differential cultural reproduction of knapping skill and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
 author: 
@@
 In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45].
-This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework.
+This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework, which also great overlaps with some core assumptions of the more modern approach of cultural transmission theory [@eerkens2005; @eerkens2007].
 Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
 Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986].
+
 For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016].
-To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
-The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]."
+A more recent approach has been to identify morphological "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
+The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]." Following this discussion, Kuhn [-@kuhnEvolutionPaleolithicTechnologies2020: 168-170] developed a complimentary framework by explicitly identifying how different factors constrain the morphology of the design target, such as production constraint (raw materials) and functional constraint (mechanical and symbolic factors).
 
@@
 
-## Skill level
+## Knapping skill
 
 Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
+At the same time, however, the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170].
 This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989].
@@
 The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by local practitioners.
-It is thus possible to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].
+It is thus possible in experimental and ethnographic settings to evaluate the skill levels reflected in knapping products [@stout2002; @roux1995].
+
 When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
@@
 We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
-As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data?
-Accordingly, we predict that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is a common target.
-2) Does training has a differential effect in terms of the reproduction of skill level and mental templates among novices?
-Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby slower than the latter aspect.
+As such, our theory-driven data-informed project has the following two interconnected research questions: 1) What can the deep structure revealed through the multivariable analysis of handaxe morphometric data inform us on the material manifestation of knapping skills and mental templates?
+Our presumption here is that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is a common target.
+2) Does training has a differential effect in terms of the reproduction of knapping skill level and mental templates among novices?
+Our expectation is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby slower than the latter aspect.
 This hypothesis is informed by the previous study of Pargeter et al.
@@
 
-Our study suggests that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the differential effects of training using the data from a 90-hour long knapping skill acquisition experiment and confirmed the previous finding [-@pargeter2019] that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.
-In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers.
+Our study suggests that skill can differentially affect the expression and of different aspects of artifact mental templates, potentially biasing processes of cultural reproduction.
+In the case of handaxe morphology, we found that skill is more highly constraining of cross-sectional thinning (PC1) than it is of handaxe elongation and pointedness (PC2).
+This is in accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], and supports the use of cross-sectional thinning as a robust indicator of skill level at Boxgrove.
+It also suggests that cultural evolutionary approaches to handaxe morphology should consider technological choices about investment in skill acquisition [@pargeter2019] as a directional influence alongside random copy error [@eerkens2005] as sources of variation.
+In contrast, we found morphological targets not requiring cross-sectional thinning (elongation and pointedness (PC2)) to be less constrained by skill.
+These aspects of morphology might thus provide a clearer signal of "arbitrary" cultural variation and accumulating copy error.
+Notably, Boxgrove handaxes are highly constrained along PC2 compared to our experimental samples, in keeping with prior arguments that production at this site adhered to of a well-defined mental template [@garcía-medrano2019; @shipton2020].
+
 Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume.
@@
 
-Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers, which supports our differential cultural reproduction hypothesis.
+Regarding the two research questions we proposed in the beginning, our case study suggested that 1) To some extent we can delineate the effects of knapping skill and mental template through the handaxe morphometric data, where the former is closely associated with cross-sectional thinning while the latter is mainly expressed in elongation and pointedness due to the constraint of the former; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers, which supports our differential cultural reproduction hypothesis.
 At a larger theoretical level it questions the distinction between social learning of design targets vs.

### 2023-01-28 manuscript/manuscript_word.Rmd
finished another round of revision.

@@
 ---
-
 # Introduction

### 2023-01-28 code/Boxgrove_Iovita.R

@@ -419,3 +419,17 @@ ggplot2::ggsave("delta weight comparison3.png", path="figure", width = 20,
 
 ggstatsplot::extract_stats(p)
 
+
+# exploring the relationship between delta weight and PC1 and PC2
+deltaw <- read.csv("data/Experiment/Weight data.csv")
+shape <- read.csv("data/Experiment/boxgrove_experiment_data_shapes.csv")
+merged = merge(shape, deltaw, by.x=c('individual', 'assessment'), by.y=c('Knapper', 'Assessment'))
+merged1 <- filter(merged, group != "expert")
+
+
+ggplot(merged1, aes(delta_Weight, PC1, color = as.factor(assessment_stage))) + 
+  geom_point()+
+  geom_smooth(method=lm)
+ggplot(merged1, aes(delta_Weight, PC2, color = as.factor(assessment_stage))) + 
+  geom_point()+
+  geom_smooth(method=lm)

### 2023-01-30 manuscript/manuscript.Rmd
incooperating Dietz's revision advice.

@@
 ---
-title: "Differential cultural reproduction of knapping skill and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
+title: "Differential effects of knapping skill on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights"
 author: 
@@
 More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017].
-From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles.
-Of particular attention here are the experimental works conducted by Stout et al.
-[-@stout2014] focusing on inferring knapping skill level and Garcia-Medrano et al.
-[-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage.
-Our paper incorporates these two perspectives into a broader conceptual framework of cultural reproduction and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.
+
+Here we used experimental data from a multidisciplinary study of handaxe-making skill acquisition [@pargeter2019; @pargeter2020; @bayani2021] to investigate the interaction between learning processes and the reproduction of specific morphological targets (c.f.
+"mental template").
+This investigation was motivated by the theoretical expectation that, just as developmental processes can act to channel biological variation and shape evolutionary trajectories [@laland2015], learning challenges might influence cultural evolutionary processes [e.g., @schillinger2014].
+Results allowed us to identify particular aspects of handaxe morphology that are more and less constrained by learning difficulty, thus helping to partition sources of morphological variation [@lycett2015].
+By comparing our experimental results to a large sample of handaxes from the site of Boxgrove, England, we were able to illustrate the use of this approach assess the presence and nature of culturally reproduced mental templates.
+This complements previous work investigated reduction intensity [@shipton2015] and raw material form [@garcía-medrano2019] as alternative explanations for morphological variation and standardization at Boxgrove.
 
@@
 
-Recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology.
-Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations.
+Current conceptions of a "mental template" are thus more nuanced than the idea of a fully specified image in the mind of the maker that is directly expressed in material form and transmitted between minds.
+For example, Hutchence and Scott [-@hutchence2021], leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations.
 From this perspective, social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning.
@@
 However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
+
+In short, contemporary approaches to the concept of a mental template emphasize the causal importance of production process and constraints and the interaction between individual and group level phenomena.
+We again note the striking similarity of the perspectives to the concept of "constructive development" as a source of guided variation in evolution biology [@laland2015].
+We thus sought to further develop these perspectives by directly investigating the effects of learning difficulty and skill acquisition on the reproduction of experimentally controlled design targets.
 

### 2023-02-01 manuscript/manuscript.Rmd
integrating more of Dietz's comments.

@@
 ---
-title: "Differential effects of knapping skill on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights"
+title: "Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights"
 author: 
@@
 We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
-Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology.
-The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the effects of training on the differential cultural reproduction of these two aspects using the data from a 90-hour-long knapping skill acquisition experiment.
-We found that the desired shape of a handaxe can be relatively quickly picked up by novices, while reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
+Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differnetial effect in the cultural reproduction of different aspects of handaxe morphology.
+More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill.
+It also confirmed that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running (90-hour) training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
 date: "`r Sys.Date()`"
@@
 In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45].
-This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework, which also great overlaps with some core assumptions of the more modern approach of cultural transmission theory [@eerkens2005; @eerkens2007].
+This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework.
 Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004].
 Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986].
+Importantly, this conception of a mental template as an idea or image transmitted between minds also echoes core assumptions of the more modern approach of cultural transmission theory [@eerkens2005; @eerkens2007].
 
-For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016].
+For a decade or so, the mental template concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignoring other competing factors such as raw material constraints [@white1995], and c) being constrained by the basic fracture mechanics and design space of bifacial technology [@moore2011; @moore2016].
 A more recent approach has been to identify morphological "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
@@
 
-The cultural reproduction, or cultural transmission as described in standard cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and skill levels makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records.
+The cultural reproduction, or transmission as it is commonly termed in the cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and production skills makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records.
 Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed.
@@
 We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
-As such, our theory-driven data-informed project has the following two interconnected research questions: 1) What can the deep structure revealed through the multivariable analysis of handaxe morphometric data inform us on the material manifestation of knapping skills and mental templates?
-Our presumption here is that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is a common target.
-2) Does training has a differential effect in terms of the reproduction of knapping skill level and mental templates among novices?
-Our expectation is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby slower than the latter aspect.
-This hypothesis is informed by the previous study of Pargeter et al.
+As such, our theory-driven data-informed project aims to examine the following research questions: 1) Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology?
+To answer this question, we first applied multivariate analysis of handaxe morphometric data to identify different components of morphological variation.
+Our study has the following three assumptions: 1) Boxgrove, expert, and novice assemblages should have a common design target (mental template) as Boxgrove is designated as the model during the training; 2) The morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences; 3) Throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby constraining the latter aspect.
+The third assumption here is oarticularly informed by the previous study of Pargeter et al.
 [-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.
@@
 
-Our study suggests that skill can differentially affect the expression and of different aspects of artifact mental templates, potentially biasing processes of cultural reproduction.
+Our study suggests that skill can differentially affect the expression of different aspects of artifact mental templates, potentially biasing processes of cultural reproduction.
 In the case of handaxe morphology, we found that skill is more highly constraining of cross-sectional thinning (PC1) than it is of handaxe elongation and pointedness (PC2).
-This is in accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], and supports the use of cross-sectional thinning as a robust indicator of skill level at Boxgrove.
-It also suggests that cultural evolutionary approaches to handaxe morphology should consider technological choices about investment in skill acquisition [@pargeter2019] as a directional influence alongside random copy error [@eerkens2005] as sources of variation.
+This is in accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], and supports the use of cross-sectional thinning as a robust indicator of skill at Boxgrove, with potential implications for the cognitive demands and social contexts supporting learning [@pargeter2019; @stout2014].
+It further suggests that cultural evolutionary approaches to handaxe morphology should consider technological choices about investments in skill acquisition [@pargeter2019] as a directional influence alongside random copy error [@eerkens2005] as sources of variation.
 In contrast, we found morphological targets not requiring cross-sectional thinning (elongation and pointedness (PC2)) to be less constrained by skill.
@@
 This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013].
-Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394], although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014].
+Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by direct [@stout2014; @bello2016] and indirect [@roberts2009; @roberts1998: 384-394] archaeological evidence from Boxgrove, although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014].
 It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages.
-In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment.
+In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, although some of them did not follow the instruction during the assessment.
 On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.
@@
 
-# Conclusions
+# Conclusion
 
-Regarding the two research questions we proposed in the beginning, our case study suggested that 1) To some extent we can delineate the effects of knapping skill and mental template through the handaxe morphometric data, where the former is closely associated with cross-sectional thinning while the latter is mainly expressed in elongation and pointedness due to the constraint of the former; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers, which supports our differential cultural reproduction hypothesis.
+Regarding the research question we proposed in the beginning, our case study suggested that skill acquisition has a differential effect in the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so.
 At a larger theoretical level it questions the distinction between social learning of design targets vs.

### 2023-02-01 manuscript/manuscript_word.Rmd
integrating more of Dietz's comments.

@@
 ---
-title: 'Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights'
+title: 'Differential effects of knapping skill acquisition on the cultural reproduction of Late Acheulean handaxe morphology: Archaeological and experimental insights'
 author:
@@
 We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
-Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology.
-The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2).
-Moreover, we also evaluated the effects of training on the differential cultural reproduction of these two aspects using the data from a 90-hour-long knapping skill acquisition experiment.
-We found that the desired shape of a handaxe can be relatively quickly picked up by novices, while reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
+Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differnetial effect in the cultural reproduction of different aspects of handaxe morphology.
+More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill.
+It also confirmed that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running (90-hour) training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
 output:
@@
 ---
+
 # Introduction

### 2023-02-05 manuscript/manuscript.Rmd
substantial revision after reframing.

@@
 abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.
-Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates.
-Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
+Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapping skill and mental templates.
+Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present new results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
 We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
-Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differnetial effect in the cultural reproduction of different aspects of handaxe morphology.
+Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differential effect in the cultural reproduction of different aspects of handaxe morphology.
 More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill.
-It also confirmed that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running (90-hour) training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
+Our findings thus shed new light on how the processes of skill learning can bias the cultural reproduction of artifact morphology.
+\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Knapping skill; Mental template; Cultural transmission"
 date: "`r Sys.Date()`"
@@
 
-The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019].
+The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019; @white2022].
 Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011; @moncel2018a; @moncel2018b; @moncel2018c].
-More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017].
+More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapping skill [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017].
 
@@
 Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
-At the same time, however, the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170].
+At the same time, however, researchers have also pointed out that the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170].
 This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989].
@@
 Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed.
-The ignorance of the latter becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a].
-This reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].
+The complexity of this issue is further exemplified by the fact that motor skills like knapping cannot be simply learned through observation but must be reconstructed through individual practice using supportive material in social contexts [@stout2017].
+The ignorance of this factor becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a].
+As we stated earlier, this reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].
 
-Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
+Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of knapping skill and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers.
 We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
-As such, our theory-driven data-informed project aims to examine the following research questions: 1) Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology?
+As such, our theory-driven data-informed project aims to examine the following research question:
+
+> Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology?
 To answer this question, we first applied multivariate analysis of handaxe morphometric data to identify different components of morphological variation.
-Our study has the following three assumptions: 1) Boxgrove, expert, and novice assemblages should have a common design target (mental template) as Boxgrove is designated as the model during the training; 2) The morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences; 3) Throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby constraining the latter aspect.
-The third assumption here is oarticularly informed by the previous study of Pargeter et al.
+Our study has the following three assumptions: 1) Boxgrove, expert, and novice assemblages should have a common design target (mental template) as Boxgrove is designated as the model before the training; 2) The morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect different level of knapping skill; 3) Throughout the training the novice samples should become more similar to expert samples in both knapping skill and mental template, but the acquisition of the former aspect will be more challenging and thereby constraining the latter aspect.
+The third assumption here is particularly informed by the previous study of Pargeter et al.
 [-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.
@@
 This 500-ka-old site has documented exceedingly rich details of Lower Paleolithic hominin subsistence behaviors [@smith2012; @smith2013] and their paleoenvironmental contexts [@holmes2010; @preece2022].
-In addition to the presence of one of the earliest hominin fossil [tentatively assigned to *Homo heidelbergensis*, @hillson2010; @lockey2022; @roberts1994] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
+In addition to the presence of one of the earliest hominin fossil [tentatively assigned to *Homo heidelbergensis*, @hillson2010; @lockey2022; @roberts1994] and bone assemblages with anthropogenic modifications in northern Europe [@bello2009], Boxgrove is mostly known for its large sample size of Late Acheulean-style flint handaxes and the high knapping skill level reflected in their manufacture (**Figure** \@ref(fig:photos)).
 As such, it has received wide research attention in the past two decades regarding the relationships between technology, cognition, and skills [@stout2014; @iovita2011; @iovita2017; @shipton2015; @garcía-medrano2019; @keyHandaxeShapeVariation2019].
-To identify the morphological manifestation of knappers' skill level in our study, we selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
+We selected a complete handaxe assemblage (n=326) previously analyzed and reported in digital formats by Iovita and McPherron [-@iovita2011], which is currently curated at the Franks House of the British Museum [@iovita2017].
 The digital photographs are taken of each handaxe at a 90$^{\circ}$ angle, which was oriented with the tip to the right of the photos, and the camera faces the most convex surface of the handaxe [@iovita2011].
@@
 
-In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas from our expert sample.
+In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas prepared by Bruce Bradley, which are part of our expert sample as described above.
 Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage.
@@
 
-Subsequently, the 17 participants in the experimental group were assessed after every ten hours of the cumulative learning period, where each of them was requested to produce a handaxe for expert knapper's (N.
+Subsequently, the 17 participants in the experimental group were assessed after every ten hours of the cumulative learning period, where each of them was requested to produce a handaxe for instructor's (N.
 Khreisheh) review, leading to the compilation of a data set composing 9 assessments in total.
 It should be also noted that 6 out of 17 participants dropped out of the research before the final assessment due to personal reasons.
-To detect the effect of training on skill level and mental template, we reorganized our assessment classification scheme and combined it into three broader categories, namely pre-training (assessment 1), early training (assessment 2-5), and late training (assessment 6-9), which helps increase the sample size of the measured intervals.
+To understand the effect of skill acquisition on artifact morphology, we reorganized our assessment classification scheme and combined it into three broader categories, namely pre-training (assessment 1), early training (assessment 2-5), and late training (assessment 6-9), which helps increase the sample size of the measured intervals.
 A more detailed experimental protocol can be assessed in one of our published papers [@pargeter2019].
@@
 Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017].
-To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test.
+To understand the process of skill learning of novices using the Boxgrove and expert samples as benchmarks, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test.
 Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test.
-The sample size of each compared group can be as low as 6 [@games1976; @sauder2019].
+The sample size of each compared group can be as low as 6 [@games1976; @sauder2019], which makes it particularly suitable for this study as the sample size of expert experimental collection is rather small.
 Lastly, we compare the delta weight, as defined by the difference between initial nodule weight and end product weight, between these groups to understand the effect of reduction intensity on morphological variation.
@@
 In addition, visual inspection of the principle component scatter plot (**Figure** \@ref(fig:GeneralPCA1)) suggested that PC1 and PC2 might be negatively correlated within the Boxgrove and Expert groups.
-To test this, we conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)).
+To test this intuition, we conducted a series of exploratory plotting and statistical analyses of the PC values of three groups analyzed in our analysis (**Figure** \@ref(fig:PCcorrelation)).
 Across all three groups, a negative correlation has been displayed between the PC1 and PC2 values, although this trend is not statistically significant (r=-0.41, p= 0.24) in the expert group, probably because of its small sample size.
@@
 
-## Effects of training
+## Effects of skill acquisition
 
 We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section.
-As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot.
-Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$).
-An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).
-These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (thinning) and PC2 (pointedness).
-However, the effects of training across different assessment periods on both dimensions are not significant.
+As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are not statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.65$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot.
+Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.33$, $p>0.05$).
+Additionally, the pairwise comparisons of mean PC2 values between the Early training and Expert ($t=-3.5$, $p>0.05$) and between the Late training and Expert ($t=-3.68$, $p>0.05$) are also not statistically significant.
+An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.82$, $p>0.05$).
+These results essentially suggest that there is a significant difference between the pre-training group and post-training groups in both PC1 (thinning) and PC2 (pointedness), while the effects of training across different assessment periods on both dimensions are not significant.
+Interestingly, the post-training groups are very different from the expert group in the mean PC1 value, but not in the mean PC2 value.
 Regarding the delta weight of different groups, our analysis (**Figure** \@ref(fig:weight)) suggests that there is a significant difference between the pre-training group and Late training group, while all other pairwise group comparison results are insignificant.
@@
 These aspects of morphology might thus provide a clearer signal of "arbitrary" cultural variation and accumulating copy error.
-Notably, Boxgrove handaxes are highly constrained along PC2 compared to our experimental samples, in keeping with prior arguments that production at this site adhered to of a well-defined mental template [@garcía-medrano2019; @shipton2020].
+Notably, Boxgrove handaxes are highly constrained along PC2 compared to our experimental samples, in keeping with prior arguments that production at this site adhered to a well-defined mental template [@garcía-medrano2019; @shipton2020].
 
@@
 This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013].
-Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by direct [@stout2014; @bello2016] and indirect [@roberts2009; @roberts1998: 384-394] archaeological evidence from Boxgrove, although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014].
+Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by direct [@stout2014; @bello2016] and indirect [@roberts2009; @roberts1998: 384-394] archaeological evidence from Boxgrove, although the validity of differentiating percussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies [@driscoll2014].
 It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages.
@@
 
-Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis.
-It should be noted that a thin cross section as measured by PC1 could also be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, however, novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter.
-Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al.
-[-@garcía-medrano2019].
-The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe." In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide an mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020].
-The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time.
+Unexpectedly, we found that modern experimental knappers did not closely approximate the PC2 (elongation and pointedness) values of Boxgrove handaxes.
+More specifically, the Boxgrove assemblage displays an ovate shape featuring a wider tip as previously pointed out by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al.
+[-@garcía-medrano2019], while the experimental assemblages are characterized by a more pointed shape with a longer central axis (**Figure** \@ref(fig:GeneralPCA1)).
+This likely reflects the fact that our expert participants and instructor were verbally requested to make handaxes "comparable to Boxgrove handaxes" (with which they were familiar), but were not provided with a concrete template to copy.
+It would appear that they thus followed a more generalized archaeological conception of a "representative teardrop Late Acheulean handaxe" with a particular focus on thinning (PC1) that likely reflects the current cultural value placed on thinning as a marker of skill [@whittaker2004: 180-182].
+Novices then sought to approximate this form, as demonstrated by their instructor and exemplar handaxes from our current expert sample that were given as models, where the mean PC2 value of exemplar handaxes reach 1.18, indicating a high degree of elongation and pointedness even among the expert sample.
+It should be also noted that the instructor (N.
+Khreisheh) has learned how to knap and how to teach knapping from one of our expert knappers (B.
+Bradley), potentially suggesting a cascading effect of social learning that also contributed to a shared mental template between the expert group and the novice group after training.
+At a higher level, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide a mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020].
+The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise by actively teaching multiple novices at the same time.
 Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions.
-Unfortunately, the handaxe data from the instructor (N.
-Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley).
-This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.
 
-The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form.
-Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes.
-In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90].
-As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182].
-It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove.
-Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers.
-This also provides an alternative explanation to the social transmission of form for the experimental convergence on pointed forms.
-In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.
-
-In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs.
-post-training) in both PC1 (skill level) and PC2 (mental template).
-Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant.
-When the performance of experts is used as a reference point here, we can see that for PC2 no significance difference is detected between early training, late training, and expert group, while for PC1 the expert group is clearly different from the training groups, supporting our hypothesis in terms of the differential cultural reproduction of mental templates and skill level.
-This finding provides a parallel line of evidence that corroborates what has been suggested in Pargeter et al.
-[-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures.
-Methodologically speaking, this study also demonstrated that the pattern revealed by the multivariate analysis of morphometric data can nicely match with the expert knapper's 5 point grading scale of novices' knapping performances that takes multiple factors into consideration, including outcome, perceptual motor execution, and strategic understanding [See Table 2 of -@pargeter2019 for more details].
-
-Moreover, this follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue.
-As previously shown in Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
-Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; 2) the present-day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019]; and/or 3) modern knappers' skill level was affected by time constraints when they were requested to produce the reference collections [@schillingerConsideringRoleTime2014; @lewis2022].
-
-The pre-training group is unexpectedly similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
+Moreover, the pre-training group is unexpectedly similar to the Boxgrove group in PC2 potentially because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
 If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113].
-This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)).
-It might be worth noting that the expert group is highly variable probably due to raw material starting size/shape.
+This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of a handaxe and the weight of its corresponding nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:weight)).
+It might be worth noting that the expert group is highly variable probably due to the raw material starting size and/or shape.
 Achieve handaxe forms while removing as little mass as possible (i.e.
@@
 
-Although we are not the first research team to use secondary archaeological data [e.g., @keyHandaxeShapeVariation2019], we would still like to highlight here that this research project further exemplifys the potential of reusing old archaeological data in digital format to address novel research questions.
-In this paper, the main source of archaeological data is a collection of photos produced and curated more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original artifacts.
-Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions.
-Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021].
-Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].
-As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository.
-Another major and legitimate concern of archaeological data reuse is their quality.
-In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic.
-Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].
+The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form.
+Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes.
+In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90].
+As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182].
+It is possible that such constraints help to explain why our novice knappers on average produced more handaxes in similar shapes to those preferred by modern expert knappers, however, this clearly does not explain the design target at Boxgrove.
+Given the ovate forms of the Boxgrove assemblage, it thus requires a high skill level to overcome this structural constraint to produce thin yet wide handaxes as demonstrated by the Boxgrove knappers.
+This also provides an alternative explanation to the cultural reproduction of form for the experimental convergence on pointed forms.
+In this comparative context, it would only be the Boxgrove assemblage that provided evidence of social conformity on a more difficult target shape.
+
+As previously shown in Key's [-@keyHandaxeShapeVariation2019] previous finding regarding Boxgrove, it is also noteworthy how constrained the range of Boxgrove assemblage morphological variation is as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied groups.
+Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; 2) the present-day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019]; and/or 3) modern knappers' skill level was affected by time constraints when they were requested to produce the reference collections [@schillingerConsideringRoleTime2014; @lewis2022].
 
@@
 
-Regarding the research question we proposed in the beginning, our case study suggested that skill acquisition has a differential effect in the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so.
-At a larger theoretical level it questions the distinction between social learning of design targets vs.
+Regarding the research question we proposed in the beginning, our case study suggested that the processes of skill learning in a lithic medium do exert biases on the cultural reproduction of artifact morphology.
+More specifically, skill acquisition has a differential effect on the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so.
+At a larger theoretical level, it questions the distinction between social learning of design targets vs.
 individual learning of the skills needed to achieve them.
-Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015].
-This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data).
-In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022].
 In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.

### 2023-02-05 code/Boxgrove_Iovita.R

@@ -188,7 +188,7 @@ pcaplot<-fviz_pca_ind(bepca1, habillage= boxgrove_experiment_data_shapes1$group,
              ellipse.type = "convex",
              legend.title = "Groups",
              label="none")
-ggsave("General PCA1.png", path="figure", dpi = 600)
+#ggsave("General PCA1.png", path="figure", dpi = 600)
 
 fviz_pca_ind(bepca1)
 library(cowplot)
@@ -338,6 +338,15 @@ p1<-ggstatsplot::ggbetweenstats(
   results.subtitle = FALSE,
   title = "A between-group comparison of PC1 values"
 )
+#highlight_bb <- boxgrove_experiment_data_shapes %>% 
+  #dplyr::filter(individual=="bb_1" | individual=="bb_2" | individual=="bb_3" | individual=="bb_4")
+#p11<-p1+  geom_point(data=highlight_bb, 
+             #aes(x=assessment_stage,y=PC1), 
+             #color='blue',
+             #size=3,
+             #alpha=0.7,
+             #shape=17)
+
 ggplot2::ggsave("PC1 comparison.png", path="figure", width = 20,
                 height = 10, units = "cm", dpi = 600)
 ggstatsplot::extract_stats(p1)
@@ -350,6 +359,12 @@ p2<-ggstatsplot::ggbetweenstats(
   results.subtitle = FALSE,
   title = "A between-group comparison of PC2 values"
 )
+#p22<-p2+  geom_point(data=highlight_bb, 
+                     #aes(x=assessment_stage,y=PC2), 
+                     #color='blue',
+                     #size=3,
+                     #alpha=0.7,
+                     #shape=17)
 ggplot2::ggsave("PC2 comparison.png", path="figure", width = 20,
                 height = 10, units = "cm", dpi = 600)
 ggstatsplot::extract_stats(p2)

### 2023-02-07 manuscript/manuscript.Rmd
ready to submit version.

@@
 Results allowed us to identify particular aspects of handaxe morphology that are more and less constrained by learning difficulty, thus helping to partition sources of morphological variation [@lycett2015].
-By comparing our experimental results to a large sample of handaxes from the site of Boxgrove, England, we were able to illustrate the use of this approach assess the presence and nature of culturally reproduced mental templates.
+By comparing our experimental results to a large sample of handaxes from the site of Boxgrove, England, we were able to illustrate the use of this approach to assess the presence and nature of culturally reproduced mental templates.
 This complements previous work investigated reduction intensity [@shipton2015] and raw material form [@garcía-medrano2019] as alternative explanations for morphological variation and standardization at Boxgrove.
@@
 A more recent approach has been to identify morphological "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].
-The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]." Following this discussion, Kuhn [-@kuhnEvolutionPaleolithicTechnologies2020: 168-170] developed a complimentary framework by explicitly identifying how different factors constrain the morphology of the design target, such as production constraint (raw materials) and functional constraint (mechanical and symbolic factors).
+The major difference between the concepts of design imperatives and mental templates lies in the fact that the former does not necessarily require the presence of explicit internal representations of form, where the shape of handaxes can instead emerge "through the coalescence of ergonomic needs in the manipulation of large cutting tools [@wynn2021: 185]." Following this discussion, Kuhn [-@kuhnEvolutionPaleolithicTechnologies2020: 168-170] developed a complementary framework by explicitly identifying how different factors constrain the morphology of the design target, such as production constraint (raw materials) and functional constraint (mechanical and symbolic factors).
 
@@
 Regarding the critique of design space constraint, Moore and Perston's experiment [-@moore2016] suggested that bifaces can be manufactured through flake removals dictated by a random algorithm.
-However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
+However, Moore [-@moore2020: 656-657] also suggested that these random experiments cannot produce "attributes like the congruent symmetries of handaxes seen in the Late Acheulean." In short, when exercised with proper caution, the concept of mental templates still has value in the study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.
 
@@
 We again note the striking similarity of the perspectives to the concept of "constructive development" as a source of guided variation in evolution biology [@laland2015].
-We thus sought to further develop these perspectives by directly investigating the effects of learning difficulty and skill acquisition on the reproduction of experimentally controlled design targets.
+We sought to further develop these perspectives by directly investigating the effects of learning difficulty and skill acquisition on the reproduction of experimentally controlled design targets.
 
@@
 Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
-At the same time, however, researchers have also pointed out that the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170].
+At the same time, however, researchers have also pointed out that the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in a recent and comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170].
 This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989].
@@
 
-When contextual information is less readily available as in the Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
-In addition to measurements that can be almost applied in any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019; @whittaker2004: 180-182] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
+When contextual information is less readily available as in Late Acheulean archaeological assemblages, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022].
+In addition to measurements that can be applied in almost any lithic technological system such as raw materials, platform preparation, as well as hinges, in the context of handaxe technology, symmetry [@hodgson2015; @hutchence2019] and cross-sectional thinning [@caruana2020; @stout2014; @pargeter2019; @whittaker2004: 180-182] have been frequently quoted as reliable and distinctive indicators of the skill level as supported by several experimental studies.
 These two features have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].
@@
 The cultural reproduction, or transmission as it is commonly termed in the cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and production skills makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records.
-Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed.
+Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015].
+Knapping skill and learning difficulties particular to the lithic medium have been less commonly considered as a potential influence of the reprodution of the form.
 The complexity of this issue is further exemplified by the fact that motor skills like knapping cannot be simply learned through observation but must be reconstructed through individual practice using supportive material in social contexts [@stout2017].
@@
 We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest.
-As such, our theory-driven data-informed project aims to examine the following research question:
-
-> Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology?
-To answer this question, we first applied multivariate analysis of handaxe morphometric data to identify different components of morphological variation.
-Our study has the following three assumptions: 1) Boxgrove, expert, and novice assemblages should have a common design target (mental template) as Boxgrove is designated as the model before the training; 2) The morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect different level of knapping skill; 3) Throughout the training the novice samples should become more similar to expert samples in both knapping skill and mental template, but the acquisition of the former aspect will be more challenging and thereby constraining the latter aspect.
-The third assumption here is particularly informed by the previous study of Pargeter et al.
-[-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.
+As such, our theory-driven data-informed project aims to examine the following research question: Do the processes of skill learning in a lithic medium exert any biases on the cultural reproduction of artifact morphology?
+To address this question, we assessed the degree to which trainees succeeded in approximating different aspects of handaxe morphology represented in a sample of modern experts and then compared both samples with archaeological handaxes from Boxgrove.
 
@@
 In the knapping skill acquisition experiment, all research participants participated in the initial assessment (assessment 1 in our data set) before formal training, where they each produced a handaxe after watching three 15-minute videos of Late Acheulean style handaxes demonstrated by expert knappers and examining four Late Acheulean style handaxe replicas prepared by Bruce Bradley, which are part of our expert sample as described above.
-Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013] with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage.
+Training was provided by verbal instruction and support from the second author, an experienced knapping instructor [@khreisheh2013], herself trained by Bruce Bradley, with 10 years knapping practice and specific knowledge of Late Acheulean technology including the Boxgrove handaxe assemblage.
 She was present at all training sessions to provide help and instruction to participants.
@@
 [-@garcía-medrano2019], while the experimental assemblages are characterized by a more pointed shape with a longer central axis (**Figure** \@ref(fig:GeneralPCA1)).
-This likely reflects the fact that our expert participants and instructor were verbally requested to make handaxes "comparable to Boxgrove handaxes" (with which they were familiar), but were not provided with a concrete template to copy.
-It would appear that they thus followed a more generalized archaeological conception of a "representative teardrop Late Acheulean handaxe" with a particular focus on thinning (PC1) that likely reflects the current cultural value placed on thinning as a marker of skill [@whittaker2004: 180-182].
-Novices then sought to approximate this form, as demonstrated by their instructor and exemplar handaxes from our current expert sample that were given as models, where the mean PC2 value of exemplar handaxes reach 1.18, indicating a high degree of elongation and pointedness even among the expert sample.
+This likely reflects the fact that our expert participants and instructor were verbally requested to make handaxes "comparable to Boxgrove handaxes" (with which they were familiar) but were not provided with a concrete template to copy.
+It would appear that they thus followed a more generalized archaeological conception of a "representative teardrop Late Acheulean handaxe" with a particular focus on thinning (PC1).
+This likely also reflects the current cultural value placed on thinning as a marker of skill [@whittaker2004: 180-182].
+Novices sought to approximate this form, as demonstrated by their instructor and exemplar handaxes.
+In fact, the four examples presented to trainees have a mean PC2 value of 1.18, indicating a high degree of elongation and pointedness.
 It should be also noted that the instructor (N.
@@
 
-Moreover, the pre-training group is unexpectedly similar to the Boxgrove group in PC2 potentially because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
+The pre-training group is unexpectedly similar to the Boxgrove group in PC2.
+This is potentially because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)).
 If the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged [@winton2005: 113].
@@
 It might be worth noting that the expert group is highly variable probably due to the raw material starting size and/or shape.
-Achieve handaxe forms while removing as little mass as possible (i.e.
+Achieving handaxe forms while removing as little mass as possible (i.e.
 making as big a handaxe as possible from the nodule) generally requires a higher skill level due to the reductive or subtractive nature of stone knapping, where correcting an error or any thinning procedure always requires the removal of raw material and thereby reducing the size of a given handaxe [@schillingerCopyingErrorCultural2014a: 130; @deetz1967: 48-49].
@@
 
-The negative correlation between the PC1 and PC2 values revealed a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form.
+The negative correlation between the PC1 and PC2 values suggests a hidden structural constraint regarding the relationship between cross-sectional thinning and the imposed form.
 Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes.
-In the thinning phase of handaxe making, a knapper must strike flakes that travel more than one half way across the surface while not breaking the handaxe into half [-@callahan1979: 90].
+In the thinning phase of handaxe making, a knapper must strike flakes that travel more than half way across the surface while not breaking the handaxe into half [-@callahan1979: 90].
 As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182].
@@
 
-Regarding the research question we proposed in the beginning, our case study suggested that the processes of skill learning in a lithic medium do exert biases on the cultural reproduction of artifact morphology.
+Our case study suggested that the processes of skill learning in a lithic medium does in fact exert biases on the cultural reproduction of artifact morphology.
 More specifically, skill acquisition has a differential effect on the cultural reproduction of different aspects of mental templates, where cross-sectional thinning (PC1) is more constrained by knapping skill while elongation and pointedness (PC2) is less so.
-At a larger theoretical level, it questions the distinction between social learning of design targets vs.
+At a larger theoretical level, these results question the distinction between social learning of design targets vs.
 individual learning of the skills needed to achieve them.

### 2023-02-07 manuscript/manuscript_word.Rmd
ready to submit version.

@@
 abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive.
-Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates.
-Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
+Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapping skill and mental templates.
+Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present new results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers.
 We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK.
-Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differnetial effect in the cultural reproduction of different aspects of handaxe morphology.
+Through the principal component analysis of morphometric data derived from images, our study suggested that knapping skill acquisition has a differential effect in the cultural reproduction of different aspects of handaxe morphology.
 More specifically, compared with elongation and pointedness (PC2), cross-sectional thinning (PC1) is more constrained by knapping skill.
-It also confirmed that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running (90-hour) training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
+Our findings thus shed new light on how the processes of skill learning can bias the cultural reproduction of artifact morphology.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
 output:

### 2023-03-15 manuscript/manuscript.Rmd
minor typos fixed.

@@
 Following the reconceptualization of the mental template as a more flexible and interactive concept, one possible way of defining skill is the capacity for a knapper to realize mental templates using the resources available [@roux1995: 66].
-At the same time, however, researchers have also pointed out that the technological choices defining a particular metal template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in a recent and comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170].
+At the same time, however, researchers have also pointed out that the technological choices defining a particular mental template may themselves be shaped by learning challenges and costs [@roux1990; @henrich2015], implying the possibility of skill development as a constraint factor on artifact form that is not highlighted even in a recent and comprehensive literature review on this topic [@kuhnEvolutionPaleolithicTechnologies2020: 168-170].
 This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989].
@@
 PC1 (Dim.1) is postively correlated with all thickness-related variables and negatively correlated with all width-related variables and the maximum length.
-PC2 (Dim.2) is positively with bottom width and thickness variables as well as the miximum length and negatively correlated with width and thickness variables of the tip area.") %>%
+PC2 (Dim.2) is positively correlated with bottom width and thickness variables as well as the miximum length and negatively correlated with width and thickness variables of the tip area.") %>%
   kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
@@
 Our results (**Figure** \@ref(fig:PCcorrelation)) suggested thinner handaxes (low PC1 value) are generally more pointed/less ovate (high PC2 value), which was first reported in Crompton and Gowlett's [-@crompton1993] pioneering study on the allometry of Kilombe handaxes.
-In the thinning phase of handaxe making, a knapper must strike flakes that travel more than half way across the surface while not breaking the handaxe into half [-@callahan1979: 90].
+In the thinning phase of handaxe making, a knapper must strike flakes that travel more than half way across the surface while not breaking the handaxe into half [@callahan1979: 90].
 As a corollary, we speculate that it would be easier to perform thinning if the plan shape of a handaxe is narrower and more pointed, echoing the high technological difficulty of making large yet thin bifacial points as perceived by American hobbyist flintknappers [@whittaker2004: 180-182].
